---
title: "dm_NCBI"
output: html_document
---

# Read in packages and functions
```{r}
source(paste0(here::here(), "/1-Config/config.R"))
```


# read in the data
```{r}
hits <- read_csv(paste0(dataDir, "closestRefInfo_allScoreThresholdHits_wSingleClosestRef_no_2018.csv")) 

metadata <- readRDS(paste0(dataDir, "/srb_lite_notes_ages_countries.rds"))
```

## the purpose of this is to remove any last non-metagenomic assemblies

# after going through the pipeline with the dataset made in removingNonAssemblies.Rmd, I downloaded one or two samples from each project that was kept (unless there were visible metagenome assemblies on NCBI)
# looked at each to see if it was contigs/scaffolds
```{r}
# this is 86 samples; they were isolate/strain assembly and one had very tiny contigs
samples_remove <- c("SAMN02674", "SAMN02675", "SAMN026761", "SAMN02676200", "SAMN02954257", "SAMN02954258", "SAMN03449989")

## remove from metadata
metadata_final <- metadata %>%
  filter(!str_detect(sample_id, paste(samples_remove, collapse = "|"))) %>%
  #remove extra replicates from PRJDB3601, PRJNA380276, and PRJNA243036
  filter(!(str_detect(sample_id, "SAMD000361|SAMD000362|SAMD000363|SAMD000364") & 
          str_detect(`sample name`, "F") &
           str_detect(`sample name`, "-"))) %>%
  filter(!(str_detect(sample_id, "SAMN08533") &
           str_detect(`sample material processing`, "tru"))) %>%
  filter(!(str_detect(notes, "systematic comparison") & 
         !str_detect(sample_id, "SAMN02712088|SAMN02712139|SAMN02712150")))

# this is 7194 samples
```

# starting with the metaHIT samples, as I have found they are likely duplicates of each other
```{r}
# all are likely the same people -> metaHIT consortium 
# these are the proper subject names based on the supp. tables for PRJEB1220, PRJEB5224, and PRJEB4336; inferred for PRJEB28245
# want to keep longitudinal data but get rid of samples that are processed in two different ways 
# keep both when they have the same subject name but end in -0, -1, -2, -3, -4 (longitudinal)
# the PT/PN samples appear to be totally different samples according to supp. table from article
metaHIT <- metadata_final %>%
  filter((str_detect(sample_id, "SAMEA233|SAMEA2041|SAMEA2042|SAMEA2043|SAMEA1972440|SAMEA1972159|SAMEA1967055|SAMEA1966253|SAMEA1965906|SAMEA1964914|SAMEA1963786|SAMEA483|SAMEA214|SAMEA215|SAMEA216|SAMEA1972381|SAMEA1972188|SAMEA1970465|SAMEA1969417|SAMEA1967891|SAMEA1965149|SAMEA1964852|SAMEA1964687"))) %>%
  mutate(`host subject id` = ifelse(str_detect(sample_id, "SAMEA233|SAMEA2041|SAMEA2042|SAMEA2043|SAMEA1972440|SAMEA1972159|SAMEA1967055|SAMEA1966253|SAMEA1965906|SAMEA1964914|SAMEA1963786"), str_extract(`sample name`, "^[^-]+"), `host subject id`)) %>%
  mutate(`host subject id` = ifelse(str_detect(sample_id, "SAMEA483"), str_extract(`sample name`, "^[^-_]+"), `host subject id`)) %>%
  mutate(`host subject id` = ifelse(str_detect(sample_id, "SAMEA214|SAMEA215|SAMEA216|SAMEA1972381|SAMEA1972188|SAMEA1970465|SAMEA1969417|SAMEA1967891|SAMEA1965149|SAMEA1964852|SAMEA1964687"), str_extract(`sample name`, "(?<=-)[^-]+"), `host subject id`)) %>%
  relocate(`sample name`, .before = `host subject id`)

# which subjects have more than one sample?
metaHIT_dups <- metaHIT %>%
  group_by(`host subject id`) %>%
  filter(n() > 1) %>%
  ungroup()

# find the sample name before _assemblies was added (PRJEB28245 samples)
prefixes <- metaHIT_dups$`sample name`[!str_detect(metaHIT_dups$`sample name`, "_assemblies")] %>% 
  str_remove("MetaHIT-")

# take the metaHIT df and remove duplicates
metaHIT_dups_removed <- metaHIT %>%
  # remove rows where sample name contains _assemblies if it is a duplicate of another sample
  filter(!(str_detect(`sample name`, "_assemblies") & 
           str_remove(`sample name`, "_assemblies") %in% prefixes)) %>%
  # remove these as they are duplicates of the V1.CD1-0-PT sample, etc.
  filter(!(str_detect(`sample name`, "V1.CD-PT1-0_assemblies|V1.CD-PT2-0_assemblies|V1.CD-PT3-0_assemblies|V1.CD-PT4-0_assemblies|V1.CD-PT6-0_assemblies")))

# the only things left over are the dups of MetaHIT-MH0001/MH0001 (others are longitudinal samples)
metaHIT_dups_left <- metaHIT_dups_removed %>%
  group_by(`host subject id`) %>%
  filter(n() > 1) %>%
  ungroup() %>%
  filter(str_detect(`sample name`, "MH"))

# which have more hits? not super critical these are 10-30 hits we are talking about
leftover1220 <- metaHIT_dups_left %>%
  filter(!(str_detect(`sample name`, "MetaHIT")))

leftover4336 <- metaHIT_dups_left %>%
  filter(str_detect(`sample name`, "MetaHIT"))

#need to go down a bit and run hits_final ages first then come back here
hits1220 <- hits %>%
  filter(biosample_id %in% leftover1220$sample_id)

hits4336 <- hits %>%
  filter(biosample_id %in% leftover4336$sample_id)

#chose to keep samples from PRJEB4336 since Bilophila hits are present and also it was published first 
metaHIT_dups_removed_final <- metaHIT_dups_removed %>%
  filter(!(sample_id %in% leftover1220$sample_id))

# get list of samples that were removed by taking difference between original metaHIT df and df with dups removed
samples_remove_metaHIT <- setdiff(metaHIT, metaHIT_dups_removed_final)

```


```{r}
metadata_final_two <- metadata_final %>%
  #853 samples to remove
  filter(!(sample_id %in% samples_remove_metaHIT$sample_id)) %>%
  # Remove extra samples from the athlete study because there was not longitudinal sampling. Keep the sample with more hits for the subject (see name in `unique ID`). For one subject, SAMN04329868 was selected since it had the same number of hits as the other sample but longer hits.
  filter(!(str_detect(sample_id, "SAMN04329816|SAMN04329819|SAMN04329764|SAMN04329873|SAMN04329767|SAMN04329773|SAMN04329774|SAMN04329778|SAMN04329779"))) %>%
  # add fixed ages for the InfantGut project since some are maternal samples (none have hits so this does not affect prevalence calculations), this is based on info found in their published article
  mutate(under_over_3 = case_when(str_detect(sample_id, "SAMEA257|SAMEA258") & str_ends(Alias, "_M") ~ "over 3", TRUE ~ under_over_3))
##6332 samples left! THIS IS THE FINAL COUNT FOR PREVALENCE CALCULATIONS
  

metadata_removed <- setdiff(metadata, metadata_final_two)
metadata_removed_one_two <- setdiff(metadata_final, metadata_final_two)

saveRDS(metadata_final_two, file = paste0(dataDir, "/metadata_final.rds"))
#0420 was the csv written before removal of duplicate metaHIT samples
#0504 was for final analysis
#0612 adds correct age bins for InfantGut study but there are no hits for any of these samples
write.csv(metadata_final_two, file = paste0(configDir, "/metadata_final_0612.csv"), row.names = FALSE)

```

```{r}
## remove from final hits 
hits_final <- hits %>%
  filter((biosample_id %in% metadata_final_two$sample_id))
#4243 hits
## add ages
hits_final_ages <- hits_final %>%
  left_join(metadata_final_two %>% select(sample_id, under_over_3), by = c("biosample_id" = "sample_id"))

saveRDS(hits_final_ages, file = paste0(dataDir, "/hits_final_ages.rds"))

```

## add subject info for samples with hits for cooccurrence network
#ie is it longitudinal? 
# move subject IDs from other columns to the host subject id and if there is none, add the biosampleID in that column (only one sample for that subject)
```{r}
metadata_final_subjects <- metadata_final_two %>%
  filter(sample_id %in% hits_final_ages$biosample_id) %>%
  # no repeats for subject sampling 
  mutate(`host subject id` = 
           ifelse(str_detect(sample_id, "SAMN1465|SAMN1238|SAMN0824|SAMN061153|SAMN061154|SAMN061155|SAMN05214|SAMN04978193|SAMN044804|SAMN04161035|SAMN02712088|SAMEA4378|SAMD00036524"), sample_id, `host subject id`)) %>%
  # only one of the subject's samples has hits 
  mutate(`host subject id` = 
           ifelse(str_detect(sample_id, "SAMN0906"), sample_id, `host subject id`)) %>%
  # no manuscript and no way to tell if samples are longitudinal
  filter(!(str_detect(sample_id, "SAMN061") & `local-scale environmental context` == "ankylosing spondylitis patient")) %>%
  # the following had patient descriptors on NCBI
  mutate(`host subject id` = 
           ifelse(str_detect(sample_id, "SAMN06018971|SAMN06018972|SAMN06018973|SAMN06018974"), "Donor", `host subject id`)) %>%
  mutate(`host subject id` = 
           ifelse(str_detect(sample_id, "SAMN06018975|SAMN06018976|SAMN06018977"), "Recipient_1", `host subject id`)) %>%
  mutate(`host subject id` = 
           ifelse(str_detect(sample_id, "SAMN06018979"), "Recipient_2", `host subject id`)) %>%
  # no repeats for subject sampling (using sample_id instead of sampleID so sample identifiers are not one or two digit numbers)
  mutate(`host subject id` = 
           ifelse(str_detect(notes, "Human gut microbiome in AS"), sample_id, `host subject id`)) %>%
  # patient descriptors on NCBI for "Human fecal microbiome before and after consumption of AH1206"
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05356889|SAMN05194733|SAMN05194711"), "Subject_X", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05356887|SAMN05194731|SAMN05194709"), "Subject_V", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05356884|SAMN05194706"), "Subject_S", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05356879|SAMN05194723|SAMN05194701"), "Subject_M", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05356872|SAMN05194694"), "Subject_E", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05194727"), "Subject_R", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05194722|SAMN05194700"), "Subject_K", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05194721|SAMN05194699"), "Subject_J", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05194720"), "Subject_I", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05194718|SAMN05194696"), "Subject_G", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05194707"), "Subject_T", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05194697"), "Subject_H", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN05194692"), "Subject_C", `host subject id`)) %>%
  # likely not from the same subject (only two samples)
  mutate(`host subject id` = 
           ifelse(str_detect(sample_id, "SAMN0502"), sample_id, `host subject id`)) %>%
  # subject IDs from "sample reference" variable; sampled at Baseline and days 28 and 56
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04158616|SAMN04158617|SAMN04158618"), "Subject_40870", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04158619"), "Subject_41952", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04158622|SAMN04158623|SAMN04158624"), "Subject_42827", `host subject id`)) %>%
  # no repeats for subject sampling
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN036536"), patient_name, `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN00715|SAMN00791|SAMN00792|SAMN00993"), sample_id, `host subject id`)) %>%
  # portion before the first dash is the subject ID (supp. table 1 from paper) PRJEB5224 (SAMEA233)
  # PRJEB1220 are the others
 mutate(`host subject id` = ifelse(str_detect(sample_id, "SAMEA233|SAMEA2041|SAMEA2042|SAMEA2043|SAMEA1972440|SAMEA1967055|SAMEA1966253|SAMEA1965906|SAMEA1964914|SAMEA1963786"), str_extract(`sample name`, "^[^-]+"), `host subject id`)) %>%
  #PRJEB28245
  mutate(`host subject id` = ifelse(str_detect(sample_id, "SAMEA483"), str_extract(`sample name`, "^[^-_]+"), `host subject id`)) %>%
  #PRJEB4336 -> participants recruited from inter99 study but have very similar naming to other metaHIT samples and project name is MetaHIT-Richness
  mutate(`host subject id` = ifelse(str_detect(sample_id, "SAMEA214|SAMEA215|SAMEA216|SAMEA1972381|SAMEA1972188|SAMEA1970465|SAMEA1969417|SAMEA1967891|SAMEA1965149|SAMEA1964852|SAMEA1964687"), str_extract(`sample name`, "(?<=-)[^-]+"), `host subject id`)) %>%
  relocate(`sample name`, .before = `host subject id`) %>%
  # keep just the samples from SAMEA273 with "individual name" since we cannot be sure about subjects
  filter(!(str_detect(sample_id, "SAMEA273") & is.na(`individual name`))) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMEA273"), `individual name`, `host subject id`)) %>%
  # the 86 duplicates were removed for this project
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMD000361|SAMD000362|SAMD000363|SAMD000364"), `sample name`, `host subject id`)) %>%
  # according to the manuscript, the subject id is the apr## portion
  mutate(`host subject id` = if_else(str_detect(`host subject id`, "apr"),
                                     str_remove(`host subject id`, "S\\d+$"),
                                     `host subject id`)) %>%
  #132 samples and 132 unique identifiers under "common name"; same for SAMEA443 with 3 samples
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMEA3708|SAMEA443"), `common name`, `host subject id`)) %>%
  # for PRJEB12123 need to remove the time points to get the subject ID (SAMEA3708)
  mutate(`host subject id` = str_remove(`host subject id`, "-S1-|-S2-|-S3-")) %>%
  # only two samples, from different countries
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN06699"), `submitted subject id`, `host subject id`)) %>%
  # based NCBI subject names for the samples poooled to make these five coassemblies
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN09866707"), "Subject_T26", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN09866710"), "Subject_T18", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN09866716"), "Subject_T16", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN09866719"), "Subject_T23", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN09866722"), "Subject_T15", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329766"), "Enve", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329776"), "SantaCruz", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329780"), "Ibis", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329851"), "Cove", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329852"), "Norco", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329853"), "SRAM", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329854"), "Thomson", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329856"), "Easton", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329857"), "Renthal", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329858"), "Jamis", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329859"), "Niner", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329860"), "Intense", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329862"), "Pivot", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329863"), "Pinarello", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329864"), "Merlin", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329865"), "Commencal", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329866"), "Yeti", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329867"), "Giant", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329868"), "Juliana", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329869"), "Huffy", `host subject id`)) %>%
  mutate(`host subject id` =
           ifelse(str_detect(sample_id, "SAMN04329871"), "Speedplay", `host subject id`))

# to double check we don't have any duplicate subject names between unrelated projects. everything looks how i expect it to 
checking <- metadata_final_subjects %>% filter(!(str_detect(`host subject id`, "SAMN|SAMD|SAMEA")))

#realized that there are a few subjects with only moorella hits. dont want them to be included when calculating percentages of SSRM-positive individuals with certain taxa
hits_final_ages_no_MA <- hits_final_ages %>% filter(!is.na(single_closest_ref)) %>% 
  filter(single_closest_ref != "Moorella_thermoacetica_Y72_copy_2__MooTher5")
# now 4097 hits

metadata_final_subjects_0710 <- metadata_final_subjects %>% 
  filter(sample_id %in% hits_final_ages_no_MA$biosample_id)
  

saveRDS(metadata_final_subjects_0710, file = paste0(dataDir, "/metadata_final_subjects.rds"))


# In this df, there are several biosamples that say "FALSE" under SRM_present but they have hits. The portions of the metadata table associated with species/hits/etc. are outdated. The only columns used in the most recent analysis of the data are the ones downloaded from NCBI.
```

# removing hits from contigs with hits to different species that could influence the cooccurrence analysis (produced after an initial look at the cooccurrence plot made using the hits_final_ages as input instead of hits_final_cooccur made below)
# i.e., if there is one contig with a hit to bilohila and a hit to d. piger it will seem like the two species are cooccurring when it is just a consequence of fragment insertion
```{r}
# which samples have hits to different species on the same contig?
contig_check <- hits_final_ages_no_MA %>%
  filter(biosample_id %in% metadata_final_subjects_0710$sample_id) %>%
  mutate(contig = str_extract(hit_id, "(?<=\\.).*(?=_)")) %>%
  group_by(biosample_id, contig, query_id) %>%
  # is there a hit to more than one species on one contig and are these part of the same cluster?
  summarize(n_diff_hits = n_distinct(single_closest_ref),
            closest_refs = paste(unique(single_closest_ref), collapse = ", ")) %>%
  filter(n_diff_hits > 1) %>%
  # remove rows where all refs are bilophila or d. piger or DsvSp (the ones in this df are from the same cluster)
  filter(!sapply(str_split(closest_refs, ", "), function(refs) all(str_detect(refs, "Bilophila")))) %>%
  filter(!sapply(str_split(closest_refs, ", "), function(refs) all(str_detect(refs, "Desulfovibrio_piger")))) %>%
  filter(!sapply(str_split(closest_refs, ", "), function(refs) all(str_detect(refs, "DsvSp")))) %>%
  # gordonibacter not an issue 
  filter(!sapply(str_split(closest_refs, ", "), function(refs) all(str_detect(refs, "Gordonibacter")))) %>%
  # these are paired with species from the same cluster
  filter(!str_detect(closest_refs, "DslDes17")) %>%
  # the NAs are not an issue because they are not in the matrix
  filter(!str_detect(closest_refs, "\\bNA\\b")) %>%
  # two uncultured bacteria from the same cluster
  filter(!str_detect(biosample_id, "SAMD00036432"))


# for the contigs in contig_check, we want to select one subunit
species_to_keep_by_subject <- hits_final_ages_no_MA %>%
  mutate(contig = str_extract(hit_id, "(?<=\\.).*(?=_)")) %>%
  #contig needs to be specific to the biosample_id
  filter(contig %in% contig_check$contig & biosample_id %in% contig_check$biosample_id) %>%
  group_by(biosample_id, contig) %>%
  #take subunit on the contig that is the closest to its reference
  slice_min(distance, with_ties = FALSE) %>%
  # remove manually extra hits from contigs that all map to NA
  # remove shorter hit
  filter(!hit_id %in% c("OPVH01.34419.1_1/63-228", 
                        "ORTY01.76057.1_1/1-124",
                        "OVYH01.71234.1_1/42-206",
                        "OWRT01.35758.1_2/1-115",
                        "OXYJ01.14281.1_2/2-102",
                        "ULWZ01.63543.1_1/1-84",
                        "UPEG01.53870.1_2/1-91",
                        "UXIV01.20473.1_2/1-126",
                        "UZHO01.27007.1_1/11-102"))

species_to_remove_by_subject <- hits_final %>%
  mutate(contig = str_extract(hit_id, "(?<=\\.).*(?=_)")) %>%
  filter(contig %in% contig_check$contig & biosample_id %in% contig_check$biosample_id) %>%
  filter(!hit_id %in% species_to_keep_by_subject$hit_id)


hits_final_cooccur <- hits_final_ages_no_MA %>%
  filter(!hit_id %in% species_to_remove_by_subject$hit_id)
#3957 hits #now 3811 after removing NA and thermoacetica

saveRDS(hits_final_cooccur, file = paste0(dataDir, "/hits_final_cooccur.rds"))
```

# also need to do this for the sample prevalence analysis; start with the dataset hits_final_cooccur and remove any other hits from one contig that are mapping to different species that show up when you include all samples
```{r}
contig_check_for_samples <- hits_final_ages %>%
  #already did these
  filter(!biosample_id %in% metadata_final_subjects$sample_id) %>%
  mutate(contig = str_extract(hit_id, "(?<=\\.).*(?=_)")) %>%
  group_by(biosample_id, contig) %>%
  # is there a hit to more than one species on one contig and are these part of the same cluster?
  summarize(n_diff_hits = n_distinct(single_closest_ref),
            closest_refs = paste(unique(single_closest_ref), collapse = ", ")) %>%
  filter(n_diff_hits > 1) %>%
  # remove rows where all refs are bilophila or d. piger
  filter(!sapply(str_split(closest_refs, ", "), function(refs) all(str_detect(refs, "Bilophila")))) %>%
  filter(!sapply(str_split(closest_refs, ", "), function(refs) all(str_detect(refs, "Desulfovibrio_piger")))) %>%
  filter(!str_detect(closest_refs, "Gordonibacter")) %>%
  # the NAs are not an issue because they are not in the matrix
  filter(!str_detect(closest_refs, "\\bNA\\b"))

species_to_keep_by_sample <- hits_final %>%
  mutate(contig = str_extract(hit_id, "(?<=\\.).*(?=_)")) %>%
  filter(contig %in% contig_check_for_samples$contig & biosample_id %in% contig_check_for_samples$biosample_id) %>%
  group_by(biosample_id, contig) %>%
  #take subunit on the contig that is the closest to its reference
  slice_min(distance, with_ties = FALSE)

species_to_remove_by_sample <- hits_final_cooccur %>%
  mutate(contig = str_extract(hit_id, "(?<=\\.).*(?=_)")) %>%
  filter(contig %in% contig_check_for_samples$contig & biosample_id %in% contig_check_for_samples$biosample_id) %>%
  filter(!hit_id %in% species_to_keep_by_sample$hit_id)

hits_final_sample_prev <- hits_final_cooccur %>%
  filter(!hit_id %in% species_to_remove_by_sample$hit_id) %>%
  # remove the one archaeoglobus hit that has a distance > 2.5 
  # the others with dist. > 2.5 (moorella thermoacetica get removed later)
  filter(!hit_id == "OOYY01.37298.1_1/16-276")

#3784 hits

saveRDS(hits_final_sample_prev, file = paste0(dataDir, "/hits_final_sample_prev.rds"))
```


## make a matrix that has a 1 for a species*subject combo when there is a hit present for that combo
# code developed with Claude AI, all double-checking done by me
```{r}
all_refs <- unique(hits_final_cooccur$single_closest_ref)
nas <- hits_final_cooccur %>% filter(is.na(single_closest_ref))
#filtering out the NAs as these appear to not have been successfully inserted into the tree. 
all_refs <- all_refs[!is.na(all_refs)]

# collapse by `host subject id`, make a sample ids variable
matrix <- metadata_final_subjects_0710 %>%
  group_by(`host subject id`) %>%
  summarise(sample_ids = paste(sample_id, collapse = ", ")) %>%
  mutate(across(all_of(all_refs), ~ 0)) # initialize all ref columns to 0

for (ref in all_refs) {
  # get all biosample_ids that have a hit for this ref
  samples_with_hit <- hits_final_cooccur$biosample_id[hits_final_cooccur$single_closest_ref == ref]
  
  # for each row, check if any of its sample_ids are in samples_with_hit
  matrix[[ref]] <- sapply(matrix$sample_ids, function(ids) {
    id_list <- str_split(ids, ", ")[[1]]
    as.integer(any(id_list %in% samples_with_hit))
  })
}

# #checking to see that I am getting the values I expect 
# 14040721 should have hits for D. piger 2,3,4 and EF645667
# Subject_42827 should have hits for D. alkalitolerans, T. narugense, Anaerobic_bacterium_sk.prop8__Bv2Prop2, and d. desulfuricans DQ092635
# apr12 should have hits for B. wadsworthia 3+6
# MH0309 should have hits for Anaerobic_bacterium_sk.prop8__Bv2Prop2
check <- matrix %>% 
  filter(str_detect(`host subject id`, "14040721|Subject_42827|apr12|MH0309"))

subject_OTU_matrix <- matrix %>% select(-"sample_ids")

saveRDS(subject_OTU_matrix, file = paste0(dataDir, "/subject_OTU_matrix_0710.rds"))


```


## make a matrix that has a 1 for a species*sample combo when there is a hit present for that combo
# code developed with Claude AI, all double-checking done by me
```{r}
all_refs_sample <- unique(hits_final_sample_prev$single_closest_ref)
#filtering out the NAs as these appear to not have been successfully inserted into the tree. 
#techincally one less OTU than the subject_OTU_matrix because the sample with the polluted aquifer clone was removed in hits_final_sample_prev (and the sample where that occurs is not part of the metadata_final_subjects df either) i.e., this is not important
all_refs_sample <- all_refs_sample[!is.na(all_refs_sample)]

matrix_sample <- metadata_final_two %>%
  filter(sample_id %in% hits_final_ages$biosample_id) %>%
  select(sample_id) %>%
  mutate(across(all_of(all_refs_sample), ~ 0)) # initialize all ref columns to 0

for (ref in all_refs_sample) {
  # place a one if a species has a hit for a given sample and a zero if not 
  samples_with_hit_samples <- hits_final_sample_prev$biosample_id[hits_final_sample_prev$single_closest_ref == ref]
  matrix_sample[[ref]] <- as.integer(matrix_sample$sample_id %in% samples_with_hit_samples)
}

# The number of samples with bilophila wadsworthia 6 and d. piger 3 is what I expect 
# checked a few samples and they have ones where they have hits 

saveRDS(matrix_sample, file = paste0(dataDir, "/sample_OTU_matrix_0506.rds"))
```
