---
title: "archive"
output: html_document
---

```{r}
## dups would get removed when placing in the tree for visualization. 
## want to make sure they aren't duplicates from different assemblies
closest_ref_no_2018 <- read_csv(paste0(dataDir, "closestRefInfo_allScoreThresholdHits_no_2018_assemblies.csv")) 
dups <- closest_ref_no_2018 %>%
  filter(is.na(distance))

## now I want to see what their corresponding dups are 
df_with_dups_list <- read_csv(paste0(dataDir, "compiled_dsrAB_scoreThreshold_assemblies_noDups_gapPercentageInfo.csv")) %>% select(id, sdups)
# this provides the "reference/first" dup that has an actual distance and all of the others are listed as that representative hit's sdups
dups$rep_id <- sapply(dups$hit_id, function(dup) {
  match_row <- df_with_dups_list %>% 
    filter(str_detect(sdups, fixed(dup)))
  if (nrow(match_row) > 0) match_row$id[1] else NA
})

#now see which dups are only mapping to rep hits that are from a different assembly
dups_from_other_assemblies <- dups %>%
  filter(sapply(seq_along(rep_id), function(i) {
    !any(str_detect(rep_id[[i]], query_id[i]))
  }))


```

# trying to figure out what went wrong with the cooccurrence network 

```{r}
# need to read in this file from clustering_original.rmd
hits_final_interest <- readRDS(file = paste0(clusterDir, "/samples_with_cooccurrence_of_interest"))

# which samples that have both d. alkalitolerans and the uncultured firmicute are missing from the final df?
cooccur_missing <- hits_final_interest %>%
  filter(!biosample_id %in% metadata_final$sample_id)
cooccur_missing_megahit <- hits_final_interest %>%
  filter(!biosample_id %in% metadata_final_two$sample_id)
cooccur_included_megahit <- hits_final_interest %>%
  filter(biosample_id %in% metadata_final_two$sample_id)
view(setdiff(cooccur_missing_megahit, cooccur_missing))
#only losing three actual samples (because they were metaHIT duplicates, e.g. they were in metadata_final but not metadata_final_two)
```

#81 samples that should have cooccurrence happening but there are only 26 showing up -> i think there is something different about the way the hits are inserted after so many were removed 
```{r}
#and then there are only 26 subjects with the two species cooccurring in the clustering.Rmd
cooccur_missing_hgm_filtered <- hgm_df_filtered %>%
  filter(sapply(str_split(sample_id, ", "), function(ids) {
    any(ids %in% hits_final_interest$biosample_id)
  }))
wut <- cooccur_missing_hgm_filtered %>% group_by(merged_subject_id) %>% summarize(n=n())

not <- setdiff(wut$merged_subject_id, subjects_with_co$.) %>% as.data.frame()

all_samples_in_hgm <- unlist(str_split(hgm_df_filtered$sample_id, ", "))

cooccur_missing_hgm_filtered <- hits_final_interest %>%
  filter(!biosample_id %in% all_samples_in_hgm)
length(unique(cooccur_missing_hgm_filtered$merged_subject_id))

# for the new subject matrix i made 
## Out of the 68 samples that are in that cooccur list, 8 were removed in this step (adding subjects) since it was not clear whether or not they have multiple samples to individuals 
# this leaves 60 samples and 43 unique subjects




```



# initial testing of making cooccurrence network
# filtering of hgm_df using metadata_final for use in clustering.smk
```{r}
hgm_df <- read_csv(paste0(configDir, "/hgm_subject_df.csv"))
samples <- metadata_final$sample_id
any_id_matches <- function(cell_value) {
  if (is.na(cell_value)) return(FALSE)
  ids <- trimws(strsplit(cell_value, ",")[[1]])
  any(ids %in% samples)
}

hgm_df_filtered <- hgm_df %>% filter(sapply(sample_id, any_id_matches))

write.csv(hgm_df_filtered, file = paste0(clusterDir, "/hgm_subject_df_filtered.csv"), row.names = FALSE)


```


# was looking at that same cooccurrence issue in the clustering.rmd
```{r}
subby <- species_subject_matrix_by_cluster[, colnames(species_subject_matrix_by_cluster) %in% not$.]
dim(otu_subject_matrix)[2]
matrix_missing_cooccurrence <- otu_subject_matrix[c("Desulfitibacter_alkalitolerans__DfiAlka2", "Anaerobic_bacterium_sk.prop8__Bv2Prop2"), ] %>% t()
sum(rowSums(matrix_missing_cooccurrence[, ]) > 1)
subjects_with_co <- rownames(matrix_missing_cooccurrence)[rowSums(matrix_missing_cooccurrence) > 1] %>% as.data.frame()
#26
hgm_df_subset <- hgm_df %>% filter(merged_subject_id %in% subjects_with_co)
```

# trying to see how many biosample_ids are associated with multiple query ids
# does having multiple query IDs mean there are duplicates in the dataset? -> you would see multiple of the same species/hits from the same sample but different assembly/sequencing run
# code generated with help from Claude
```{r}
mult_queries <- srb_lite_notes_ages_countries$sample_id[sapply(strsplit(srb_lite_notes_ages_countries$query_ids, ","), length) > 1]

#extract the sample_id and query_ids for samples with multiple queries
samples_w_mult_queries <- srb_lite_notes_ages_countries %>% filter(sample_id %in% mult_queries) %>% select(sample_id, query_ids)

queries_long <- samples_w_mult_queries %>%
  mutate(query_ids = str_extract_all(query_ids, "[A-Z0-9]+")) %>%  # parse the string lists
  unnest(query_ids)

hits_for_assemblies <- read_csv(paste0(dataDir, "closestRefInfo_allScoreThresholdHits_no_2018_assemblies.csv"))

# are there samples that have hits from more than one different query?
samples_w_multiple_query_hits <- hits_for_assemblies %>% filter(query_id %in% queries_long$query_ids) %>%
  rename(query_ids = query_id) %>%
  left_join(queries_long, by = "query_ids") %>%
  # don't need to worry about the ones hitting moorella thermoacetica
  filter(distance < 2.5 | is.na(distance)) %>%
  group_by(sample_id) %>%
  summarise(n_query_ids = n_distinct(query_ids)) %>%
  filter(n_query_ids > 1)

#now, find which hits these are -> which samples with multiple query IDs actually had both hit something
just_filtered <- hits_for_assemblies %>% filter(query_id %in% queries_long$query_ids) %>%
  rename(query_ids = query_id) %>%
  left_join(queries_long, by = "query_ids") %>%
  filter(sample_id %in% samples_w_multiple_query_hits$sample_id) %>%
  group_by(sample_id) %>%
  summarise(query_list = list(unique(query_ids)))

# make df for .txt file of these queries that can be used with snakemake rule
qoi <- just_filtered %>%
  mutate(query_list = str_extract_all(query_list, "[A-Z0-9]+")) %>%
  unnest(query_list)

# queries that both have hits and are from one sample
write(qoi$query_list, file = "queries_dups_04-15-2026.txt")

# there are 59 biosamples that have several queries that both have hits
# all of these look like viable assemblies aside from 2 samples. 
# instead select by the query with more hits ??
which_query <- hits_for_assemblies %>%
  filter(query_id %in% qoi$query_list)

## instead of trying to choose them below, I can do this after removing other duplicates... left_join the sample ID then group_by
# SAMD00036196 OQGT01 (the other has 4 genes on one contig)
# SAMD00036214 OQHG01 (higher bit score for the hit to the same species)
# SAMD00036228 OPYT01 (the other has 3 genes on one contig)
# SAMD00036229 OQCB01 (shorter distance to the same species)
# SAMD00036251 OQAD01 (had a slightly higher bit score for a hit to the same species)
# SAMD00036316 OQEN01 (the other only had one species aside from moorella with a bit score barely over 100)
# SAMD00036330 OQDZ01 (higher bit score for same species)
# SAMD00036337 OQDT01 (3 hits to one contig for bilophila species as opposed to one)
# SAMD00036377 OQNO01 (much longer length hits)
# SAMD00036381 OQLH01 (much longer length hits)
# SAMD00036386 OQNI01 (subunits are mapping to same species)
# SAMD00036430 OQLI01 (more non-sdup hits)
# SAMD00036432 OQNV01 (more species)
# SAMEA1963786 OPRH01 (longer hits)



```

#wanted to see if I actually need to make the matrix in python; i dont 
```{r}
# 14040721 should have hits for D. piger 2,3,4 and EF645667
# 71869016 should have hits for moorella thermoacetica
# D124_G should have hits for D. piger 3 and 4
# D113 should have hits for Anaerobic_bacterium_sk.prop8__Bv2Prop2, B. wadsworthia 6, D. alkalitolerans, D. vulgaris, Gordonibacter_pamelaeae__GrdPame5, and EF645667
# D205_M should have hits for B. wadsworthia 6, D. piger 3, D. vulgaris, and EF645667
otu_subject_df <- read_csv(paste0(clusterDir, "/otu_subject_matrix_ref_0420.csv")) %>% select(c("...1","14040721","71869016","D124_G","D113","D205_M"))
# these all match except for the hits where distance = NA (meaning it is an sdup). That is fine because we still want to include those!
# just keep in mind that the reference is populated from whatever sequence the sdup is assigned to 
```

---
title: "clustering_original"
output: html_document
date: "2026-04-30"
---
# purpose is to replicate figure 2 from the original manuscript using RC's hgm_subject_df -> make sure code was working 

# Read in packages and functions
```{r}
source(paste0(here::here(), "/1-Config/config.R"))
```

# Read in OTU table and taxa table
```{r}
taxa_table <- read_csv(paste0(clusterDir, "/taxa_table_0420.csv"))
otu_subject_df_not_filtered <- read_csv(paste0(clusterDir, "/otu_subject_matrix_ref_0430_notFiltered.csv"))
otu_subject_matrix <- otu_subject_df_not_filtered %>%
  column_to_rownames(., var = "...1") %>%
  as.matrix()
```
# Read in tree and root
```{r}
library(phyloseq)
ref_tree <- read.tree(paste0(raxmlDir, "/RAxML_originalLabelledTree_noBootstrap_no_2018.newick"))
rooted_ref_tree <- root(ref_tree, c("Moorella_thermoacetica_Y72_copy_2__MooTher5"), r = TRUE)
TREE <- phy_tree(rooted_ref_tree)
```

# prep reference phyloseq object
```{r}
OTU = otu_table(otu_subject_matrix, taxa_are_rows = TRUE)

# format taxa_table
taxa_table <- taxa_table %>%
  column_to_rownames(var = "OTU") %>%
  select(-c("environmental_category","dsrab_type"))

taxa_table <- as.matrix(taxa_table)
# missing two taxa in the clusters_df (see code at bottom) so I am renaming them in taxa_table to match their names in the other phyloseq components. DslDes16 and DsvAlas8 have the same taxonomical classifications
rownames(taxa_table) <- dplyr::recode(rownames(taxa_table),
  "Candidatus_Desulfovibrio_trichonymphae__AB444072" = "Candidatus_Desulfovibrio_trichonymphae",
  "Desulfovibrio_desulfuricans_G20__DslDes16" = "Desulfovibrio_desulfuricans__DslDes16_DsvAlas8"
)
TAX = tax_table(taxa_table)

# make phyloseq object
phyloseq <- phyloseq(TREE, TAX, OTU)
```


# tip_glom at 0.15 and get list of OTUs that are in each cluster
```{r}
ps_tipglom <- tip_glom(phyloseq, h = 0.15)
# find which tips were retained after tip_glom
cluster_reps <- taxa_names(ps_tipglom)

# find the distances between tips on the tree so that each OTU can be assigned to its closest relative retained after tip_glom
tree_dist <- cophenetic(phy_tree(phyloseq))
```

# Read in info on closest references to hits
```{r}
closestRefInfo_allScoreThresholdHits_no_2018_noFilter <- read_csv(paste0(treeInfoDir, "closestRefInfo_allScoreThresholdHits_no_2018.csv"))

get_single_ref <- function(ref_list) {
  ref_list <- gsub("\\]", "", ref_list)
  ref_list <- gsub("\\[", "", ref_list)
  ref_list <- gsub(" ", "", ref_list)
  ref_list <- gsub("'", "", ref_list)
  ref_list <- strsplit(ref_list, ",")[[1]]
  
  # check if all of the closest ref sequences are Anantharaman2018 seqs
  if (length(grep("QUERY", ref_list)) == length(ref_list)) {
    # single ref is assigned as first seq in list
    return(ref_list[1])
  }
  
  # case if there is a non-Anantharaman2018 seq in the list of closest_ref
  if (length(grep("QUERY", ref_list)) != length(ref_list)) {
    single_ref_index <- grep("QUERY", ref_list, invert = T)[[1]]
    return(ref_list[single_ref_index])
  }
}

#non filtered NCBI 
closestRefInfo_allScoreThresholdHits_wSingleClosestRef_no_2018_noFilter <- closestRefInfo_allScoreThresholdHits_no_2018_noFilter %>%
  mutate(single_closest_ref = sapply(closest_ref, get_single_ref))
```



# make df that gives the cluster representative for each of the 1256 taxa in the non tip-glommed ps object
```{r}
clusters_df <- data.frame(
  OTU = taxa_names(phyloseq),
  cluster = ifelse(taxa_names(phyloseq) %in% cluster_reps,
                   taxa_names(phyloseq),
                   sapply(taxa_names(phyloseq), function(tip) {
                     cluster_reps[which.min(tree_dist[tip, cluster_reps])]
                   }))
)

# read in hits_final

# want to include only the OTUs that have hits 
names <- closestRefInfo_allScoreThresholdHits_wSingleClosestRef_no_2018_noFilter$single_closest_ref
hits_in_cluster_df <- clusters_df %>% filter(OTU %in% names) 

# testing in console
for (cluster in unique(hits_in_cluster_df$cluster)) {
  otus_in_cluster <- hits_in_cluster_df$OTU[hits_in_cluster_df$cluster == cluster]
  print(paste("Cluster:", cluster))
  print(otus_in_cluster)
  print(paste("Length:", length(otus_in_cluster)))
}

# # write one file per cluster
for (cluster in hits_in_cluster_df$cluster) {
  otus_in_cluster <- hits_in_cluster_df$OTU[hits_in_cluster_df$cluster == cluster]
  if (length(otus_in_cluster) > 1) {
    cluster_clean <- gsub("ö", "o", cluster)
    cluster_clean <- gsub("sp-", "sp.", cluster_clean)
    cluster_clean <- gsub("sk-", "sk.", cluster_clean)
    write_lines(otus_in_cluster,
                file = paste0(clusterDir, "/", cluster_clean, "_otus_in_cluster_015.txt"))
  }
}
```

```{r}
## missing two taxon at the end
## there are some OTUs from the reference tree fasta that are named differently than the reference taxonomy fasta (presumably they are two species/taxa that are very close)
ps_object <- unique(hits_in_cluster_df$OTU)
hits_final_cluster <- closestRefInfo_allScoreThresholdHits_wSingleClosestRef_no_2018_noFilter
hits_cluster <- unique(hits_final_cluster$single_closest_ref)
hits_not_ps_object <- setdiff(hits_cluster, ps_object)
#moorella was removed and the NAs have "FALSE" for srm_present
# need to make sure DslDes16, DsvAlas8, and Candidatus Desulfovibrio triconymphae are kept

# figure out where they are missing and change the names before creating ps object
taxon <- "Candidatus_Desulfovibrio_trichonymphae"

# Check presence in each component
taxon %in% phy_tree(TREE)$tip.label        # in tree?
taxon %in% rownames(taxa_table)            # in TAX?
taxon %in% rownames(otu_subject_matrix)    # in OTU?

taxon2 <- "Desulfovibrio_desulfuricans__DslDes16_DsvAlas8"

# Check presence in each component
taxon2 %in% phy_tree(TREE)$tip.label        # in tree?
taxon2 %in% rownames(taxa_table)            # in TAX?
taxon2 %in% rownames(otu_subject_matrix)    # in OTU?
```



# rename the OTUs in the matrix to species-level based on the clustering 
```{r}
otu_subject_matrix_subset <- otu_subject_matrix %>% .[rowSums(.) > 0, ] 
species_subject_matrix <- otu_subject_matrix %>% .[rowSums(.) > 0, ] 
species_subject_matrix <- species_subject_matrix[rownames(species_subject_matrix) 
                                                 != "Moorella_thermoacetica_Y72_copy_2__MooTher5", ]

rownames(species_subject_matrix) <- hits_in_cluster_df$cluster[match(rownames(species_subject_matrix), 
                                                                     hits_in_cluster_df$OTU)]
row_names <- rownames(otu_subject_matrix_subset)
# seeing if the OTUs from the matrix match the OTUs from the clustering
setdiff(row_names, hits_in_cluster_df$OTU)
# only moorella thermoacetica is missing which is what I expect since that is used for rooting the tree
setdiff(hits_in_cluster_df$OTU, row_names)
# why are these two (Desulfitobacterium_hafniense_Y51__DsfHaf15 and Bioreactor_clone__UncS2028) not in the species_subject_matrix but they are in the clustering?
# SAMN02299420 and SAMEA6784066 are the only two samples with hits, respectively, and they are not in the matrix (not in hgm_subject_df that makes the matrix, either)

# double check the row sums are the same before and after renaming 
# uncultured bacterium EF645667 -> 175 samples with a hit
rowSums(otu_subject_matrix_subset[2:6, ])
rowSums(species_subject_matrix[1:5, ])
# B. wadsworthia -> 625 samples with a hit
rowSums(otu_subject_matrix_subset[19:20, ])
rowSums(species_subject_matrix[18:19, ])
# D. piger 3 -> 285 samples with a hit
rowSums(otu_subject_matrix_subset[16:18, ])
rowSums(species_subject_matrix[15:17, ])
```
## collapse the rows by cluster and make the cooccurrence network
```{r}
species_subject_matrix_by_cluster <- rowsum(species_subject_matrix,
                                           gsub("\\.[1-4]$", "", rownames(species_subject_matrix)))

# again, make sure sums make sense
# these match the total sums found for the three clusters in the previous chunk
rowSums(species_subject_matrix_by_cluster[1:27, ])
```

```{r}
cooccur.subjectCluster <- cooccur(mat=species_subject_matrix_by_cluster, spp_names = T)
edgeList <- cooccur.subjectCluster$results[, c("sp1_name", "sp2_name", "obs_cooccur")]
edgeList <- subset(edgeList, obs_cooccur > 20)

igraph_net <- graph_from_data_frame(d = edgeList, directed = F)
E(igraph_net)$width <- E(igraph_net)$obs_cooccur/10
V(igraph_net)$label.size <- 0
deg <- igraph::degree(igraph_net, mode="all")
V(igraph_net)$size <- 1


radian.rescale <- function(x, start=0, direction=1) {
  c.rotate <- function(x) (x + start) %% (2 * pi) * direction
  c.rotate(scales::rescale(x, c(0, 2 * pi), range(x)))
}
lab.locs <- radian.rescale(x=1:vcount(igraph_net), direction=-1, start=0)

l <- layout_in_circle(igraph_net)

plot(igraph_net, layout = l, edge.curved=.3, vertex.label = NA, edge.alpha = 0.5, vertex.color = "black", vertex.frame.color = "black")


plot(igraph_net, layout = l, edge.curved=.3, vertex.color = "black", vertex.frame.color = "black", vertex.label.cex = 0.7, vertex.label.color = "black", vertex.label.degree = lab.locs, vertex.label.dist=4, edge.alpha = 0.5)

```
# which samples have both the anaerobic bacterium and d. alkalitolerans?
```{r}
interest <- c("Anaerobic_bacterium_sk.prop8__Bv2Prop2", "Desulfitibacter_alkalitolerans__DfiAlka2")
hits_final_interest <- hits_final_cluster %>% 
  filter(single_closest_ref %in% interest) %>%
  group_by(biosample_id) %>%
  summarize(n_species = n_distinct(single_closest_ref)) %>%
  filter(n_species == length(interest))

saveRDS(hits_final_interest, file = paste0(clusterDir, "/samples_with_cooccurrence_of_interest"))
```


