---
title: "archive"
output: html_document
---

```{r}
## dups would get removed when placing in the tree for visualization. 
## want to make sure they aren't duplicates from different assemblies
closest_ref_no_2018 <- read_csv(paste0(dataDir, "closestRefInfo_allScoreThresholdHits_no_2018_assemblies.csv")) 
dups <- closest_ref_no_2018 %>%
  filter(is.na(distance))

## now I want to see what their corresponding dups are 
df_with_dups_list <- read_csv(paste0(dataDir, "compiled_dsrAB_scoreThreshold_assemblies_noDups_gapPercentageInfo.csv")) %>% select(id, sdups)
# this provides the "reference/first" dup that has an actual distance and all of the others are listed as that representative hit's sdups
dups$rep_id <- sapply(dups$hit_id, function(dup) {
  match_row <- df_with_dups_list %>% 
    filter(str_detect(sdups, fixed(dup)))
  if (nrow(match_row) > 0) match_row$id[1] else NA
})

#now see which dups are only mapping to rep hits that are from a different assembly
dups_from_other_assemblies <- dups %>%
  filter(sapply(seq_along(rep_id), function(i) {
    !any(str_detect(rep_id[[i]], query_id[i]))
  }))


```

# trying to figure out what went wrong with the cooccurrence network 

```{r}
# need to read in this file from clustering_original.rmd
hits_final_interest <- readRDS(file = paste0(clusterDir, "/samples_with_cooccurrence_of_interest"))

# which samples that have both d. alkalitolerans and the uncultured firmicute are missing from the final df?
cooccur_missing <- hits_final_interest %>%
  filter(!biosample_id %in% metadata_final$sample_id)
cooccur_missing_megahit <- hits_final_interest %>%
  filter(!biosample_id %in% metadata_final_two$sample_id)
cooccur_included_megahit <- hits_final_interest %>%
  filter(biosample_id %in% metadata_final_two$sample_id)
view(setdiff(cooccur_missing_megahit, cooccur_missing))
#only losing three actual samples (because they were metaHIT duplicates, e.g. they were in metadata_final but not metadata_final_two)
```

#81 samples that should have cooccurrence happening but there are only 26 showing up -> i think there is something different about the way the hits are inserted after so many were removed 
```{r}
#and then there are only 26 subjects with the two species cooccurring in the clustering.Rmd
cooccur_missing_hgm_filtered <- hgm_df_filtered %>%
  filter(sapply(str_split(sample_id, ", "), function(ids) {
    any(ids %in% hits_final_interest$biosample_id)
  }))
wut <- cooccur_missing_hgm_filtered %>% group_by(merged_subject_id) %>% summarize(n=n())

not <- setdiff(wut$merged_subject_id, subjects_with_co$.) %>% as.data.frame()

all_samples_in_hgm <- unlist(str_split(hgm_df_filtered$sample_id, ", "))

cooccur_missing_hgm_filtered <- hits_final_interest %>%
  filter(!biosample_id %in% all_samples_in_hgm)
length(unique(cooccur_missing_hgm_filtered$merged_subject_id))

# for the new subject matrix i made 
## Out of the 68 samples that are in that cooccur list, 8 were removed in this step (adding subjects) since it was not clear whether or not they have multiple samples to individuals 
# this leaves 60 samples and 43 unique subjects




```



# initial testing of making cooccurrence network
# filtering of hgm_df using metadata_final for use in clustering.smk
```{r}
hgm_df <- read_csv(paste0(configDir, "/hgm_subject_df.csv"))
samples <- metadata_final$sample_id
any_id_matches <- function(cell_value) {
  if (is.na(cell_value)) return(FALSE)
  ids <- trimws(strsplit(cell_value, ",")[[1]])
  any(ids %in% samples)
}

hgm_df_filtered <- hgm_df %>% filter(sapply(sample_id, any_id_matches))

write.csv(hgm_df_filtered, file = paste0(clusterDir, "/hgm_subject_df_filtered.csv"), row.names = FALSE)


```


# was looking at that same cooccurrence issue in the clustering.rmd
```{r}
subby <- species_subject_matrix_by_cluster[, colnames(species_subject_matrix_by_cluster) %in% not$.]
dim(otu_subject_matrix)[2]
matrix_missing_cooccurrence <- otu_subject_matrix[c("Desulfitibacter_alkalitolerans__DfiAlka2", "Anaerobic_bacterium_sk.prop8__Bv2Prop2"), ] %>% t()
sum(rowSums(matrix_missing_cooccurrence[, ]) > 1)
subjects_with_co <- rownames(matrix_missing_cooccurrence)[rowSums(matrix_missing_cooccurrence) > 1] %>% as.data.frame()
#26
hgm_df_subset <- hgm_df %>% filter(merged_subject_id %in% subjects_with_co)
```

# trying to see how many biosample_ids are associated with multiple query ids
# does having multiple query IDs mean there are duplicates in the dataset? -> you would see multiple of the same species/hits from the same sample but different assembly/sequencing run
# code generated with help from Claude
```{r}
mult_queries <- srb_lite_notes_ages_countries$sample_id[sapply(strsplit(srb_lite_notes_ages_countries$query_ids, ","), length) > 1]

#extract the sample_id and query_ids for samples with multiple queries
samples_w_mult_queries <- srb_lite_notes_ages_countries %>% filter(sample_id %in% mult_queries) %>% select(sample_id, query_ids)

queries_long <- samples_w_mult_queries %>%
  mutate(query_ids = str_extract_all(query_ids, "[A-Z0-9]+")) %>%  # parse the string lists
  unnest(query_ids)

hits_for_assemblies <- read_csv(paste0(dataDir, "closestRefInfo_allScoreThresholdHits_no_2018_assemblies.csv"))

# are there samples that have hits from more than one different query?
samples_w_multiple_query_hits <- hits_for_assemblies %>% filter(query_id %in% queries_long$query_ids) %>%
  rename(query_ids = query_id) %>%
  left_join(queries_long, by = "query_ids") %>%
  # don't need to worry about the ones hitting moorella thermoacetica
  filter(distance < 2.5 | is.na(distance)) %>%
  group_by(sample_id) %>%
  summarise(n_query_ids = n_distinct(query_ids)) %>%
  filter(n_query_ids > 1)

#now, find which hits these are -> which samples with multiple query IDs actually had both hit something
just_filtered <- hits_for_assemblies %>% filter(query_id %in% queries_long$query_ids) %>%
  rename(query_ids = query_id) %>%
  left_join(queries_long, by = "query_ids") %>%
  filter(sample_id %in% samples_w_multiple_query_hits$sample_id) %>%
  group_by(sample_id) %>%
  summarise(query_list = list(unique(query_ids)))

# make df for .txt file of these queries that can be used with snakemake rule
qoi <- just_filtered %>%
  mutate(query_list = str_extract_all(query_list, "[A-Z0-9]+")) %>%
  unnest(query_list)

# queries that both have hits and are from one sample
write(qoi$query_list, file = "queries_dups_04-15-2026.txt")

# there are 59 biosamples that have several queries that both have hits
# all of these look like viable assemblies aside from 2 samples. 
# instead select by the query with more hits ??
which_query <- hits_for_assemblies %>%
  filter(query_id %in% qoi$query_list)

## instead of trying to choose them below, I can do this after removing other duplicates... left_join the sample ID then group_by
# SAMD00036196 OQGT01 (the other has 4 genes on one contig)
# SAMD00036214 OQHG01 (higher bit score for the hit to the same species)
# SAMD00036228 OPYT01 (the other has 3 genes on one contig)
# SAMD00036229 OQCB01 (shorter distance to the same species)
# SAMD00036251 OQAD01 (had a slightly higher bit score for a hit to the same species)
# SAMD00036316 OQEN01 (the other only had one species aside from moorella with a bit score barely over 100)
# SAMD00036330 OQDZ01 (higher bit score for same species)
# SAMD00036337 OQDT01 (3 hits to one contig for bilophila species as opposed to one)
# SAMD00036377 OQNO01 (much longer length hits)
# SAMD00036381 OQLH01 (much longer length hits)
# SAMD00036386 OQNI01 (subunits are mapping to same species)
# SAMD00036430 OQLI01 (more non-sdup hits)
# SAMD00036432 OQNV01 (more species)
# SAMEA1963786 OPRH01 (longer hits)



```



