---
title: "clustering"
output: html_document
Author: "Elizabeth Daigle"
date: "16 March 2026"
---
# Read in packages and functions
```{r}
source(paste0(here::here(), "/1-Config/config.R"))
```

# Read in OTU table and taxa table
```{r}
taxa_table <- read_csv("taxa_table.csv")
otu_subject_df <- read_csv(paste0(clusterDir, "/otu_subject_matrix_ref_0316.csv"))
otu_subject_matrix <- otu_subject_df %>%
  column_to_rownames(., var = "...1") %>%
  as.matrix()
```
# Read in tree and root
```{r}
ref_tree <- read.tree(paste0(raxmlDir, "/RAxML_originalLabelledTree_noBootstrap_no_2018.newick"))
rooted_ref_tree <- root(ref_tree, c("Moorella_thermoacetica_Y72_copy_2__MooTher5"), r = TRUE)
TREE <- phy_tree(rooted_ref_tree)
```
## HERE 

# prep reference phyloseq object
```{r}
OTU = otu_table(otu_subject_matrix, taxa_are_rows = TRUE)

# format taxa_table
taxa_table <- taxa_table %>% select(-"...1") %>%
  column_to_rownames(var = "OTU") %>%
  select(-c("environmental_category","dsrab_type"))

taxa_table <- as.matrix(taxa_table)
TAX = tax_table(taxa_table)

# make phyloseq object
phyloseq <- phyloseq(TREE, TAX, OTU)

## want to add sample data?
```

# tip_glom at 0.1 and get list of OTUs that are in each cluster
```{r}
ps_tipglom <- tip_glom(phyloseq, h = 0.1)
# find which tips were retained after tip_glom
cluster_reps <- taxa_names(ps_tipglom)

# find the distances between tips on the tree so that each OTU can be assigned to its closest relative retained after tip_glom
tree_dist <- cophenetic(phy_tree(phyloseq))
```

```{r}
clusters_df <- data.frame(
  OTU = taxa_names(phyloseq),
  cluster = ifelse(taxa_names(phyloseq) %in% cluster_reps,
                   taxa_names(phyloseq),
                   sapply(taxa_names(phyloseq), function(tip) {
                     cluster_reps[which.min(tree_dist[tip, cluster_reps])]
                   }))
)

# # write one file per cluster
# for (cluster in cluster_reps) {
#   otus_in_cluster <- clusters_df$OTU[clusters_df$cluster == cluster]
#   if (length(otus_in_cluster) > 1) {
#     write_lines(otus_in_cluster,
#                 file = paste0(clusterDir, "/", cluster, "_otus_in_cluster.txt"))
#   }
# }
```


## from ani_zoom_trees.Rmd by RC

# overall purpose: get a table for each reference leaf and which cluster they fall into with h=0.1 for tip_glom

# packages to install
```{r}
library(ggplot2)
library(treeio)
library(ggtree)
library(dplyr)
library(ape)
library(phytools)
library(phyloseq)
```

# prep tree, OTU, and TAX
```{r}
pruned_tree <- drop.tip(rooted_ref_tree, c(moorella_copy_2_labels, ox_bact_labels))

hit_otus <- unique(subset(closestRefInfo_allScoreThresholdHits_singleRefOnly$closest_single_ref, closestRefInfo_allScoreThresholdHits_singleRefOnly$distance < 2.5 & !is.na(closestRefInfo_allScoreThresholdHits_singleRefOnly$closest_single_ref)))
hit_otus <- gsub("[.]", "-", hit_otus)
hit_otus <- gsub("sp-", "sp.", hit_otus)
hit_otus <- gsub("sk-", "sk.", hit_otus)
hit_otus <- gsub("str-", "str.", hit_otus)
```

```{r}
faca_tax <- subset_taxa(TAX, supercluster == "FACA group sensu lato")

faca_tree <- phyloseq(faca_tax, OTU, pruned_tree)

plot_tree(tip_glom(faca_tree, h = 0), label.tips = "taxa_names" %in% hit_otus, text.size = 2)
plot_tree(tip_glom(faca_tree, h = 0.25), label.tips = "taxa_names", text.size = 2)
```

# make df for each leaf in each cluster at h = 0.2
```{r}
df <- taxa_table_firmicutes_notipglom

df$tip_glom_01_clusters <- ifelse(df$otu %in% df$tip_glom_01_clusters,
                              df$otu,
                              NA)

for (n_row in 1:nrow(df)) {
  if (is.na(df[n_row, "tip_glom_01_clusters"])) {
    df[n_row, "tip_glom_01_clusters"] <- df[n_row - 1, "tip_glom_01_clusters"]
  }
}

clusters_freq <- as.data.frame(table(df$tip_glom_01_clusters))
clusters <- subset(clusters_freq$Var1, clusters_freq$Freq > 1)

taxa_table_firmicutes_notipglom <- df

# for (cluster in clusters) {
#   otus_in_cluster <- df$out[df$tip_glom_01_clusters == cluster]
# 
#   write_lines(otus_in_cluster, file = paste("/Users/rebeccachristensen/Desktop/Cremer_Lab_2022/main/workflow/out/cluster_otus/ids/firmicutes/", cluster, "_otus_in_cluster.txt", sep = ""))
# }

# write_lines(unique(clusters),
#             file = "/Users/rebeccachristensen/Desktop/Cremer_Lab_2022/main/config/firmicutes_cluster_names.txt", 
#             sep = "\n")

```


