---
title: "removingNonAssemblies"
output: html_document
author: "Elizabeth Daigle and Jessica Grembi"
---
## the purpose of this script is to remove all of the NCBI samples that are not true metagenomic assemblies since they skew our calculation of SSRM prevalence
# Read in packages, functions, and directories
```{r}
source(paste0(here::here(), "/1-Config/config.R"))
```
# read in the dataframe and have it default to read values as characters so there are not parsing issues
# i don't care if it is numeric or not because i just want to make this readable for Denise and Liam

```{r}

srb <- read_csv(paste0(dataDir, "dataset21and22_withNayfachData.csv"), col_types = cols(.default = "c"))

srb_viz <- srb %>%
  select(-1) %>%
  select(-c(Bilophila_wadsworthia__BilWads6:Desulfovibrio_carbinoliphilus__JQ031012)) %>%
  select(-c(Desulfovibrio_gigas__DsvGiga4:Desulfovibrio_gabonensis__DslGabo2)) %>%
  select(-c(uncultured_bacterium__EF645665:`Rice_field_soil_clone_Bu-23__JN615148`)) %>%
  # move any project, subject, age identifiers to the front
  relocate(host, `host age`, subject_id, `host subject id`, `individual name`, age, age_years, `host_age [months]`, age_nayfach, age_group, age_years_nayfach, ProjectAccession, `project name`, .after = sample_id)
```

```{r}
toss <- c("whole genome analysis of diarrheagenic Escherichia coli isolated in Kawasaki, Japan", "Molecular epidemiology of mcr-1-positive Enterobacteriaceae", "Molecular epidemiology of carbapenem-resistant Enterobacteriaceae", "Genomic analysis of non-sorbitol-fermenting and beta-glucuronidase-positive O157:H7 strain", "Molecular characterization of IMP-1 producing Enterobacter cloacae complex isolates in multi tertiary care hospital in Tokyo, Japan","Bifidobacterium strains genome sequencing", "whole genome sequence of macrolide resistant Campylobacter jejuni", "Draft Genome Sequences of Bifidobacterium species Consecutively Isolated from Healthy Japanese.", "Whole genome sequences of motile and non-motile strains of Escherichia albertii", "Bifidobacterium pseudocatenulatum NBRC 113353 genome sequencing project", "Draft genome sequencing of Escherichia albertii", "WGS of E. coli", "Genomic analysis of the vanA-plasmid transmission among vancomycin-resistant Enterococcus in a nosocomial outbreak in Osaka, Japan", "Draft whole-genome sequencing of IMP-producing Enterobacter cloacae complex", "STEC O157:H7 intrafamily infection", "Helicobacter pylori type 4 secretion system within the integrative and conjugative element and cag pathogenicity island are gastroduodenal disease markers", "Genome-wide association study with machine-learning between H. pylori gastric-cancer and duodenal-ulcer strain reveal discriminatory amino acid differences and novel oncoprotein candidates", "Genome-wide association study", "Draft genome sequencing of Adlercreutzia equolifaciens strain 9CBH6", "Molecular epidemiology of carbapenem-resistant Enterobacteriaceae isolated in Japan", "Salmonella_Montevideo_outbreak","Molecular epidemiology of carbapenem-resistant Enterobacteriaceae isolated in Japan" )
```
```{r}
srb_lite <- srb_viz %>% 
  filter(host != "Bacteria" | is.na(host)) %>% 
  filter(!`project name` %in% toss) %>% 
  filter(`investigation type` == "metagenome" | is.na(`investigation type`)) %>% 
  filter(is.na(`organism modifier note`)) %>% 
  filter(is.na(`biosample of physical sample`)) %>%  ##Only isolates/MAGs have included the biosample.of.physical.sample
  filter(!quality_assessment_method %in% c("CheckM")) %>%  #Thees are all isolates
  filter(!`broad-scale environmental context` %in% c("bodily fluid material biome [ENVO:02000019]","anthropogenic terrestrial biome")) %>%  ## These are all isolates
  filter(!`relationship to oxygen` %in% c("anaerobe", "facultative", "obligate anaerobe")) %>% # These are all isolates
  filter(!`isolation source` %in% c("10 mM sulfoquinovose-amended vegetarian human fecal microcosm","stool isolate", "stool culture", "Tissue from the tricuspid valve", "Vulva exsudate", "stomach", "river water", "Lung tissue", "Material from the bottom lip", "Material from the ethmoid", "Bone, muscular tissue, and connective tissue from the spine extension of third thoracic vertebra","arsenic-impacted human gut samples taken from 10 individuals over 4 consecutive days", "Material from the bottom lip", "Material from the ethmoid", "Bone, muscular tissue, and connective tissue from the spine extension of third thoracic vertebra","arsenic-impacted human gut samples taken from 10 individuals over 4 consecutive days")) %>% # These are bacterial genomes isolated from stool cultures or collected from non-stool locations
  filter((!grepl("iopsy", `isolation source`) & !grepl("fluid", `isolation source`) & !grepl("chromium", `isolation source`)) | is.na(`isolation source`)) %>% #~400 stomach and gastric biopsies plus a few skin biopsies and 2 peritoneal fluid samples
  filter(`geographic location` != "USA: University of Michigan Hospital, Ann Arbor, MI" | is.na(`geographic location`)) %>% ## These sample are MAGs assembled from a metagenome
  filter(!sample_id %in% c("SAMN11080843","SAMN11080845")) %>% # these are cultures
  filter(!str_detect(sample_id, "SAMN1957")) %>% # these are individual bacteria
  filter(!str_detect(sample_id, "SAMN0729")) %>% # these are individual bacteria
  filter(!str_detect(sample_id, "SAMN0998")) %>% # these are individual bacteria or pathogens
  filter(!str_detect(sample_id, "SAMEA7690")) %>% # these are isolates
  filter(!str_detect(sample_id, "SAMEA1042")) %>% # single organism
  filter(!str_detect(sample_id, "SAMEA5175")) %>% # these are isolates
  filter(!str_detect(sample_id, "SAMN1796")) %>% # these are MAGs or pathogens
  filter(!str_detect(sample_id, "SAMD00114")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMD0015")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA1031")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA205")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA206")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA207")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA208")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA217")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA299")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA310")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA346")) %>% # isolates
  filter(!str_detect(sample_id, "SAMEA3491|SAMEA3493")) %>% # not from human stool; cultured bacteria or RNA (see VE Wagner papers)
  filter(!str_detect(sample_id, "SAMEA3496|SAMEA352")) %>% # isolates
  filter(!str_detect(sample_id, "SAMEA361")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA3677")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA3715|SAMEA3735|SAMEA3935|SAMEA436")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA4444|SAMEA4614|SAMEA4645|SAMEA4697|SAMEA4741|SAMEA4775|SAMEA4852")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA4969|SAMEA5555|SAMEA5556|SAMEA5842|SAMEA5843")) %>% #isolates
  filter(!str_detect(sample_id, "SAMEA559")) %>% #WGS of ecoli
  filter(!str_detect(sample_id, "SAMEA581|SAMEA58492|SAMEA58493|SAMEA5848|SAMEA58494|SAMEA6082")) %>% #c. diff
  # Unified Human Gastrointestinal Genome (UHGG) collection, a resource combining 286,997 genomes representing 4,644 prokaryotic species from the human gut.
  filter(!str_detect(sample_id, "SAMEA58497|SAMEA58498|SAMEA58499|SAMEA585")) %>% # single organisms
  filter(!str_detect(sample_id, "SAMEA6084|SAMEA6792|SAMEA6813|SAMEA6921|SAMEA7281|SAMEA7280|SAMEA750|SAMEA7540|SAMEA766|SAMEA788|SAMEA813|SAMEA860|SAMEA894|SAMEA899")) %>% # single organisms; c. diff; e. coli; isolates
  filter(!str_detect(sample_id, "SAMN0000|SAMN0069|SAMN0086|SAMN009902|SAMN009903|SAMN009915|SAMN00998|SAMN0111|SAMN0104|SAMN01773|SAMN01774")) %>% # "Generic sample from unclassified Bacteria/sequence"
  filter(!str_detect(sample_id, "SAMN0001|SAMN0003|SAMN0011|SAMN00262|SAMN00249|SAMN00222|SAMN00210|SAMN0062|SAMN0061|SAMN0071510|SAMN00717|SAMN0072|SAMN0073|SAMN0075|SAMN00765|SAMN00767|SAMN00769|SAMN00779|SAMN00760|SAMN0098|SAMN009918|SAMN0105|SAMN0108")) %>% # individual organisms
  filter(!str_detect(sample_id, "SAMN0082|SAMN0085")) %>% # e. coli
  filter(!str_detect(sample_id, "SAMN01779|SAMN0178|SAMN0179")) %>% # "generic sample..." or pathogen or clinical sample
  filter(!str_detect(sample_id, "SAMN018|SAMN019|SAMN020447|SAMN0205|SAMN0220|SAMN021|SAMN0221|SAMN0222|SAMN0226|SAMN02314|SAMN02317|SAMN023245|SAMN02344|SAMN0235")) %>% # "generic sample..." or individual org. or pathogen; SAMN0189.. is a biopsy sample
  filter(!str_detect(sample_id, "SAMEA215029|SAMEA215048")) %>%# individual species
  filter(!str_detect(sample_id, "SAMN02363|SAMN02368|SAMN0240|SAMN0241|SAMN0243|SAMN0246|SAMN0247|SAMN0259|SAMN0263|SAMN0264|SAMN0266|SAMN02673|SAMN0267628|SAMN0267629|SAMN026763|SAMN0268|SAMN0269|SAMN027113|SAMN02713|SAMN02719|SAMN0272|SAMN0273|SAMN0274|SAMN0276|SAMN0277|SAMN0278|SAMN0279")) %>% #individual species
  filter(!str_detect(sample_id, "SAMN02802|SAMN02803|SAMN02808|SAMN02837|SAMN0284|SAMN0285|SAMN0286|SAMN0287|SAMN0288|SAMN0289")) %>% #individual species
  filter(!(str_detect(sample_id, "SAMN029") & !str_detect(sample_id, "SAMN02954"))) # 029 is individual species but want to keep four within this string that have assemblies!
##this should have 54,654 samples

srb_lite_two <- srb_lite %>%
  filter(!str_detect(sample_id, "SAMN030|SAMN031|SAMN033|SAMN035|SAMN038")) %>% # all individual species
  filter(!(str_detect(sample_id, "SAMN032") & 
           !(age == "1 month" | 
             str_detect(sample_id, "SAMN03283") | 
             str_detect(sample_id, "SAMN032958")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN034") & 
           !(str_detect(sample_id, "SAMN034182") | 
             str_detect(sample_id, "SAMN03449989")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN036") & 
           !(str_detect(sample_id, "SAMN036536")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN037") & 
           !(str_detect(sample_id, "SAMN0377735")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN039") & 
           !(str_detect(sample_id, "SAMN0392")))) %>% # potential metagenomes
  filter(!str_detect(sample_id, "SAMN040")) %>% # all individual species
  filter(!(str_detect(sample_id, "SAMN041") & 
           !(str_detect(sample_id, "SAMN041586|SAMN04168|SAMN041610")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN042") & 
           !(`environmental medium` == "feces"))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN043") & 
           !(`broad-scale environmental context` == "human-associated habitat"))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN044") & 
           !(str_detect(sample_id, "SAMN044140|SAMN04436|SAMN044804")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN045") & 
           !(`broad-scale environmental context` == "human gut"))) %>% # potential metagenomes
  filter(!str_detect(sample_id, "SAMN046|SAMN047|SAMN048")) %>% # all individual species
  filter(!(str_detect(sample_id, "SAMN049") & 
           !(str_detect(sample_id, "SAMN04978193")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN050") & 
           !(str_detect(sample_id, "SAMN0502|SAMN0506")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN051|SAMN052") & 
           !(str_detect(sample_id, "SAMN05194|SAMN05214")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN053") & 
           !(str_detect(`isolation source`, "gut") | 
             str_detect(sample_id, "SAMN0535")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN054|SAMN055|SAMN056|SAMN057") & 
           !(str_detect(sample_id, "SAMN055785|SAMN05780991")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN058") & 
           !(`broad-scale environmental context` == "ENVO:2100002"))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN059"))) %>% # individual species
  filter(!(str_detect(sample_id, "SAMN060") & 
           !(`geographic location` %in% c("USA: Chicago","USA: Cincinnati, Ohio") | `local-scale environmental context` == "ankylosing spondylitis patient"))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN061") & 
           !(`local-scale environmental context` == "ankylosing spondylitis patient"|str_detect(sample_id, "SAMN061153|SAMN061154|SAMN061155")|`nucleic_acid` == "dna"))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN062") & 
           !(`geographic location` == "USA: virginia"))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN063|SAMN064|SAMN065|SAMN066|SAMN067|SAMN068|SAMN069") & 
           !(str_detect(sample_id, "SAMN0634|SAMN066993")))) %>% # potential metagenomes
  # double check there are not things in here that we want 
  filter(!`isolation source` %in% c("perirectal swab", "Perirectal swab", "Perirectal Swab", "rectal swab", "rectal Swab", "Rectal swab", "Rectal Swab", "RECTAL SWAB", "rectal swab neonate", "rectum", "Rectum", "rectal", "Rectal", "RECTAL")) %>% # we don't want rectal samples or saliva samples
  filter(!`environmental medium` %in% c("saliva")) %>% # don't want these two saliva samples
  mutate(country = str_split_fixed(`geographic location`, ":", n = 2)[, 1]) 
## this should have 41882 samples
 
```
```{r}
# 23,339 samples
srb_lite_three <- srb_lite_two %>%
  filter(!(str_detect(sample_id, "SAMN071|SAMN072|SAMN073|SAMN074|SAMN075|SAMN076|SAMN077|SAMN078|SAMN079") & 
           !(str_detect(sample_id, "SAMN07501|SAMN079777")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN080|SAMN081|SAMN082|SAMN083|SAMN084|SAMN085|SAMN086|SAMN087|SAMN088|SAMN089") & 
           !(str_detect(sample_id, "SAMN0824|SAMN08533")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN090|SAMN092|SAMN093|SAMN094|SAMN095|SAMN096|SAMN097|SAMN098|SAMN099") & 
           !(str_detect(sample_id, "SAMN09062|SAMN09711|SAMN09866")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN10|SAMN11|SAMN12|SAMN13|SAMN14") & 
           !(str_detect(sample_id, "SAMN10720|SAMN1100|SAMN1238|SAMN1465")))) %>% # potential metagenomes
  filter(!(str_detect(sample_id, "SAMN00771448"))) %>% # "generic sample from unclassified sequences"
  filter(!(str_detect(sample_id, "SAMN02324081"))) %>% #metagenomic library from human feces
  filter(!(str_detect(sample_id, "SAMN02382"))) %>% # individual species
  filter(!(str_detect(sample_id, "SAMEA214|SAMEA215|SAMEA216") & 
           (!is.na(host) & host == "Homo sapiens"))) %>% # ind species 
  filter(!(str_detect(sample_id, "SAMN0634"))) %>% #duplicate of other AS study
  filter(!(str_detect(sample_id, "SAMN062"))) %>% #RNAseq
  filter(!(str_detect(sample_id, "SAMEA253"))) #duplicate download of MGWAS T2D data

#7,389 samples
srb_lite_four <- srb_lite_three %>%
    filter(!(str_detect(sample_id, "SAMN15|SAMN16|SAMN17|SAMN18|SAMN19|SAMN2") & 
              !(str_detect(sample_id, "SAMN17071939|SAMN17103336")))) %>%
    filter(!(str_detect(sample_id, "SAMN02954256"))) %>% # there is already another sample for this project and there is only one subject; this one is likely a MAG
    filter(!(str_detect(sample_id, "SAMEA104427653|SAMEA104407483"))) %>% #ind species
    filter(!(str_detect(sample_id, "SAMEA1318"))) %>% # c.diff
    filter(!(str_detect(sample_id, "SAMEA1966600|SAMEA1966754|SAMEA1967299"))) %>% # ind species
    filter(!(str_detect(sample_id, "SAMEA1971427|SAMEA1972019|SAMEA1972044|SAMEA1972962"))) %>% # ind species
    filter(!(str_detect(sample_id, "SAMEA3497897"))) %>% #this is an assembly, but it is DNA isolated from virus-like particles (VLPs) recovered from fecal samples
    filter(!(str_detect(sample_id, "SAMN02044"))) # sequencing of viral particles purified from stool


srb_lite_four_wnotes <- srb_lite_four %>%  
  mutate(country = case_when(
    grepl("SAMD000365", sample_id) ~ "Japan",
    TRUE ~ country),
    age_years = case_when(
      sample_id == "SAMD00036524" ~ 30,
      sample_id == "SAMD00036525" ~ 28,
      sample_id == "SAMD00036526" ~ 0.75,
      sample_id == "SAMD00036527" ~ 37,
      sample_id == "SAMD00036528" ~ 36,
      sample_id == "SAMD00036529" ~ 3,
      sample_id == "SAMD00036530" ~ 1.5,
      sample_id == "SAMD00036531" ~ 45,
      sample_id == "SAMD00036532" ~ 0.5,
      sample_id == "SAMD00036533" ~ 35,
      sample_id == "SAMD00036534" ~ 0.5,
      sample_id == "SAMD00036535" ~ 0.5,
      sample_id == "SAMD00036536" ~ 24,
      grepl("collected", `isolation source`) ~ 0.1, # These are from Matt Olm's NEC infant cohort (Olm MR et al., "Necrotizing enterocolitis is preceded by increased gut bacterial replication, Klebsiella, and fimbriae-encoding bacteria.", Sci Adv, 2019 Dec;5(12):eaax5727)
    ),
    notes = case_when(
      `geographic location` %in% c("Finland:Espoo", "Estonia:Tartu", "Russia:Petrozavodsk") ~ "metagenomes from children in the DIABIMMUNE cohort, followed from birth to age 3",
      `project name` == "EKmeta" ~ "metagenome fecal microbiota from diet study",
      `project name` == "Infant" ~ "InfantGut: Dynamics and Stabilization of the Human Gut Microbiome during the First Year of Life",
      str_detect(sample_id, "SAMEA103958") ~ "comparing fecal microbiomes of L-DOPA-naïve PD subjects to controls by metagenomic analysis",
      str_detect(sample_id, "SAMEA104411163") ~ "infant gut microbiome study",
      str_detect(sample_id, "SAMEA2041|SAMEA2042|SAMEA2043|SAMEA1963|SAMEA1964348|SAMEA1964895|SAMEA1964914|SAMEA1965033|SAMEA1965112|SAMEA1965191|SAMEA1965384|SAMEA1965767|SAMEA1965852|SAMEA1965906|SAMEA1966253|SAMEA1967055|SAMEA1968035|SAMEA1968151|SAMEA1968361|SAMEA1969296|SAMEA1971886|SAMEA1972159|SAMEA1972229|SAMEA1972440|SAMEA1972463|SAMEA1972794|SAMEA1973207") ~ "method for identifying metagenomic species and variable genetic elements by exhaustive co-abundance binning",
      str_detect(sample_id, "SAMEA214|SAMEA215|SAMEA216|SAMEA1964270|SAMEA19646|SAMEA1964852|SAMEA1965149|SAMEA1965858|SAMEA1966821|SAMEA1967319|SAMEA1967891|SAMEA1969126|SAMEA1969178|SAMEA1969417|SAMEA1970465|SAMEA1970582|SAMEA1970858|SAMEA1970897|SAMEA1971210|SAMEA1972002|SAMEA1972188|SAMEA1972381") ~ "Richness of human gut microbiome correlates with metabolic markers",
      str_detect(sample_id, "SAMEA233") ~ "human gut metagenome from stool samples; integrated catalog of reference genes in the human gut microbiome",
      str_detect(sample_id, "SAMEA273") ~ "gut microbiome dysbiosis in RA",
      str_detect(sample_id, "SAMEA3136|SAMEA317") ~  "MGWAS on stools from healthy controls, advanced adenoma and carcinoma patients",
      str_detect(sample_id, "SAMEA336") ~  "study examining microbiota after cholera infection",
      str_detect(sample_id, "SAMEA354") ~ "metagenome-wide association study on fecal samples from 74 CRC patients and 54 controls from China",
      str_detect(sample_id, "SAMEA3708") ~ "metagenome-wide association studies on fecal samples: gut microbial dysbiosis in young adults with obesity",
      str_detect(sample_id, "SAMEA3879") ~ "stool shotgun metagenomics from study looking at association between gut microbiome and CRC",
      str_detect(sample_id, "SAMEA3882") ~ "shotgun sequencing after FMT",
      str_detect(sample_id, "SAMEA3888") ~ "changes in taxonomy and functional potential after Roux-en-Y gastric bypass",
      str_detect(sample_id, "SAMEA4370|SAMEA4384") ~ "metagenomic sequencing of the microbiota obtained from faecal samples from 6 healthy humans; investigating role of spore-forming bacteria",
      str_detect(sample_id, "SAMEA4378") ~ "identifying microbial metabolites associated with coronary heart disease; stool samples were collected in addition to the urine and plasma samples",
      str_detect(sample_id, "SAMEA4431") ~ "metagenomic shotgun sequencing; impact of EEN on Crohn's disease microbiome",
      str_detect(sample_id, "SAMEA483") ~ "Assemblies of 760 MetaHIT metagenomes used in a pangenomic study",
      str_detect(sample_id, "SAMN00265") ~ "shotgun metagenomic data from stool samples collected from four pairs of matched T1D case-control subjects",
      str_detect(sample_id, "SAMN00715|SAMN00791|SAMN00792|SAMN00993") ~ "A metagenome-wide association study of gut microbiota in type 2 diabetes",
      str_detect(sample_id, "SAMN0084") ~ "gut microbiota of an morbidly obese human",
      str_detect(sample_id, "SAMN009901") ~ "Metagenomic analysis of infant stool sample; a premature infant born at 28 weeks gestation",
      str_detect(sample_id, "SAMN0176") ~ "Symptomatic atherosclerosis gut metagenome",
      str_detect(sample_id, "SAMN02316") ~ "3rd week of life of a premature male infant; metagenome",
      str_detect(sample_id, "SAMN02364") ~ "Gut metagenome of a premature infant isolated from fecal samples",
      str_detect(sample_id, "SAMN02674|SAMN02675") ~ "human gut metagenome, microbiome study of the RISK cohort",
      str_detect(sample_id, "SAMN026761|SAMN02676200") ~ "human gut metagenome, C.diff FMT",
      str_detect(sample_id, "SAMN027112|SAMN02712") ~ "systematic comparison of the DNA extraction methods on feacal samples; gut metagenome under data type and organism",
      str_detect(sample_id, "SAMN02954257|SAMN02954258") ~ "metagenome sample from human gut metagenome",
      str_detect(sample_id, "SAMN0295426") ~ "A metagenomic study of the microorganisms in the human distal gut",
      (str_detect(sample_id, "SAMN032") & (age == "1 month")) ~ "Fecal microbiome of eczema children",
      str_detect(sample_id, "SAMN03283") ~ "Human gut microbiome metagenome, gut microbiome diversity",
      str_detect(sample_id, "SAMN032958|SAMN079777") ~ "Metagenomes from human infant fecal samples with and without necrotizing enterocolitis",
      str_detect(sample_id, "SAMN034182") ~ "Metagenome sequencing of the Hadza hunter-gatherer gut microbiota",
      str_detect(sample_id, "SAMN03449989") ~ "Homo sapiens fecal metagenome Raw sequence reads",
      str_detect(sample_id, "SAMN036536") ~ "Gut microbiome profiling of pediatric patients undergoing stem cell transplantation",
      str_detect(sample_id, "SAMN0377735") ~ "Evaluation of vertical transmission between mother-infant",
      str_detect(sample_id, "SAMN0392") ~ "Characterization of the distal gut microbiota in obese patients following a weight-loss intervention using whole metagenome shotgun sequencing",
      str_detect(sample_id, "SAMN041586") ~ "Functional dynamics of the elderly gut microbiome during probiotic consumption",
      str_detect(sample_id, "SAMN04168") ~ "This project investigates how library preparation and sequencing protocol influence the taxonomy and function analysis of microbial species in human microbiome research",
      str_detect(sample_id, "SAMN041610") ~ "Metagenomes from 11 human infant fecal samples hospitalized in the same ICU",
      (str_detect(sample_id, "SAMN04209") & (`environmental medium` == "feces")) ~ "Gut microbiome diversity among the Cheyenne & Arapaho of Oklahoma",
      (str_detect(sample_id, "SAMN043") & (`broad-scale environmental context` == "human-associated habitat")) ~ "Metagenomic sequencing to identify and measure the abundances of microorganisms present in stool samples from athletes",
      str_detect(sample_id, "SAMN044140") ~ "Human Faecal Samples Raw sequence reads",
      str_detect(sample_id, "SAMN04436") ~ "Metagenomic sequencing of preterm infant gut microbiota raw sequence reads",
      str_detect(sample_id, "SAMN044804") ~ "prospective cohort study using metagenomics to characterize gut microbiota populations of pediatric short bowel syndrome",
      (str_detect(sample_id, "SAMN04571") & (`broad-scale environmental context` == "human gut")) ~ "Human Gut Microbiome in a Multiplex Family Study of Type 1 Diabetes Mellitus",
      str_detect(sample_id, "SAMN04978193") ~ "human gut and groundwater harbor non-photosynthetic bacteria; fecal sample 3 from healthy adult C",
      str_detect(sample_id, "SAMN0502") ~ "We applied shotgun metagenomics to stool samples collected from two geographically isolated (Alabama and Colorado) outbreaks, where the etiologic agents were identified as distinct strains of Salmonella enterica serovar Heidelberg by culture-dependent methods.",
      str_detect(sample_id, "SAMN0506") ~ "Increased intestinal microbial diversity following fecal microbiota transplant for active Crohn's disease",
      str_detect(sample_id, "SAMN05194|SAMN0535") ~ "Human fecal microbiome before and after consumption of AH1206 - Raw sequence reads",
      str_detect(sample_id, "SAMN05214") ~ "Gut metagenomes were isolated from people living in the Central region of Russia",
      str_detect(sample_id, "SAMN053014|SAMN053015") ~ "Raw reads of the microbiota of premature infant gut",
      str_detect(sample_id, "SAMN055785") ~ "Metagenomics analysis of donor and recepient samples from fecal transplant for Recurrent Clostridium difficile infection",
      str_detect(sample_id, "SAMN05780991") ~ "Infant gut microbiome Metagenomic assembly",
      (str_detect(sample_id, "SAMN058") & (`broad-scale environmental context` == "ENVO:2100002")) ~ "longitudinal analysis of the developing gut microbiome in infants from Finland, Estonia, and Russian Karelia",
      (str_detect(sample_id, "SAMN060") & (`local-scale environmental context` == "ankylosing spondylitis patient")) ~ "Human gut microbiome in AS",
      (str_detect(sample_id, "SAMN060") & (`geographic location` == "USA: Chicago")) ~ "De novo reconstruction of bacterial genomes allows high-resolution tracking of microbial colonization in Fecal Microbiota Transplantation experiments",
      (str_detect(sample_id, "SAMN060") & (`geographic location` == "USA: Cincinnati, Ohio")) ~ "Gut microbiome profiling of pediatric patients undergoing stem cell transplantation",
      (str_detect(sample_id, "SAMN061") & (`local-scale environmental context` == "ankylosing spondylitis patient")) ~ "human gut metagenome in RA; this does not make sense since it says ankylosing spondylitis patient",
      str_detect(sample_id, "SAMN061153|SAMN061154|SAMN061155") ~ "Metagenomic sequencing; this study presented the bacterial diversity of human stool sample from the Vogt-Koyanagi-Harada disease patients in Chinese population",
      (str_detect(sample_id, "SAMN061") & (`nucleic_acid` == "dna")) ~ "Human gut metagenome raw sequence reads",
      str_detect(sample_id, "SAMN066993") ~ "comprehensive metagenomic study of the fecal microbiomes from 23 TD patients and seven healthy travelers",
      str_detect(sample_id, "SAMN07501") ~ "preterm infant gut metagenomes",
      str_detect(sample_id, "SAMN0824") ~ "Human gut microbiome in IBD",
      str_detect(sample_id, "SAMN08533") ~ "Human gut metagenome, marine sediment metagenome, and synthetic metagenome; our samples are human metagenome assemblies",
      str_detect(sample_id, "SAMN09062") ~ "Metagenome or environmental sample from human gut metagenome, assemblies included",
      str_detect(sample_id, "SAMN09711") ~ "Human metagenome isolated from feces of healthy adult male",
      str_detect(sample_id, "SAMN09866") ~ "Temporal and Spatial Variation of the Human Microbiota During Pregnancy",
      str_detect(sample_id, "SAMN10720|SAMN1238") ~ "nanopore long read assembly of the human gut metagenome",
      str_detect(sample_id, "SAMN1100") ~ "Resistant microbiome underlies antibiotic therapy in pouchitis",
      str_detect(sample_id, "SAMN1465") ~ "Proximity ligation based characterization of 10 human gut samples",
      str_detect(sample_id, "SAMD000361|SAMD000362|SAMD000363|SAMD000364") ~ "metagenomics of Japanese gut microbiomes",
      str_detect(sample_id, "SAMD000365") ~ "Gut metagenomes collected from 13 healthy individuals",
      str_detect(sample_id, "SAMN17071939|SAMN17103336") ~ "human gut metagenome sequencing and assembly",
      TRUE ~ "None"
    )) %>%
    relocate(notes, .after = `project name`)


#this was exporting a csv for Liam to look at 
#write.csv(srb_lite, file = "srb_lite_20-03.csv", row.names = FALSE)
```

## adding ages
```{r}
srb_lite_notes_ages <- srb_lite_four_wnotes %>%
  mutate(under_over_3 = case_when(
    str_detect(sample_id, "SAMEA103958") ~ "over 3",
    str_detect(sample_id, "SAMEA104411163") ~ "under 3",
    str_detect(sample_id, "SAMEA2041|SAMEA2042|SAMEA2043|SAMEA1963|SAMEA1964348|SAMEA1964895|SAMEA1964914|SAMEA1965033|SAMEA1965112|SAMEA1965191|SAMEA1965384|SAMEA1965767|SAMEA1965852|SAMEA1965906|SAMEA1966253|SAMEA1967055|SAMEA1968035|SAMEA1968151|SAMEA1968361|SAMEA1969296|SAMEA1971886|SAMEA1972159|SAMEA1972229|SAMEA1972440|SAMEA1972463|SAMEA1972794|SAMEA1973207") ~ "over 3",
    str_detect(sample_id, "SAMEA215|SAMEA1964270|SAMEA19646|SAMEA1964852|SAMEA1965149|SAMEA1965858|SAMEA1966821|SAMEA1967319|SAMEA1967891|SAMEA1969126|SAMEA1969178|SAMEA1969417|SAMEA1970465|SAMEA1970582|SAMEA1970858|SAMEA1970897|SAMEA1971210|SAMEA1972002|SAMEA1972188|SAMEA1972381") ~ "over 3",
    str_detect(sample_id, "SAMEA233") ~ "over 3",
    str_detect(sample_id, "SAMEA273") ~ "over 3",
    str_detect(sample_id, "SAMEA214") ~ "over 3",
    str_detect(sample_id, "SAMEA216") ~ "over 3",
    str_detect(sample_id, "SAMEA245") ~ "over 3",
    str_detect(sample_id, "SAMEA257|SAMEA258") ~ "under 3",
    str_detect(sample_id, "SAMEA3136|SAMEA317") ~ "over 3",
    str_detect(sample_id, "SAMEA354") ~ "over 3",
    str_detect(sample_id, "SAMEA3708") ~ "over 3",
    str_detect(sample_id, "SAMEA3879") ~ "over 3",
    str_detect(sample_id, "SAMEA3882") ~ "over 3",
    str_detect(sample_id, "SAMEA3888") ~ "over 3",
    str_detect(sample_id, "SAMEA4378") ~ "over 3",
    str_detect(sample_id, "SAMEA4431") ~ "over 3",
    str_detect(sample_id, "SAMEA483") ~ "over 3",
    str_detect(sample_id, "SAMN00265") ~ "under 3",
    str_detect(sample_id, "SAMN00715|SAMN00791|SAMN00792|SAMN00993") ~ "over 3",
    str_detect(sample_id, "SAMN0084") ~ "over 3",
    str_detect(sample_id, "SAMN009901") ~ "under 3",
    str_detect(sample_id, "SAMN0176") ~ "over 3",
    str_detect(sample_id, "SAMN02316") ~ "under 3",
    str_detect(sample_id, "SAMN02364") ~ "under 3",
    str_detect(sample_id, "SAMN02674") ~ "over 3",
    str_detect(sample_id, "SAMN02675") ~ "over 3",
    str_detect(sample_id, "SAMN026761") ~ "over 3",
    str_detect(sample_id, "SAMN02676200") ~ "over 3",
    str_detect(sample_id, "SAMN027112") ~ "over 3",
    str_detect(sample_id, "SAMN02712") ~ "over 3",
    str_detect(sample_id, "SAMN02954257|SAMN02954258") ~ "over 3",
    str_detect(sample_id, "SAMN0295426") ~ "over 3",
    (str_detect(sample_id, "SAMN032") & (age == "1 month")) ~ "under 3",
    str_detect(sample_id, "SAMN032958") ~ "under 3",
    str_detect(sample_id, "SAMN034182") ~ "over 3",
    str_detect(sample_id, "SAMN0377735") ~ "under 3",
    str_detect(sample_id, "SAMN0392") ~ "over 3",
    str_detect(sample_id, "SAMN041586") ~ "over 3",
    str_detect(sample_id, "SAMN04168") ~ "over 3",
    str_detect(sample_id, "SAMN041610") ~ "under 3",
    (str_detect(sample_id, "SAMN04209") & (`environmental medium` == "feces")) ~ "over 3",
    (str_detect(sample_id, "SAMN043") & (`broad-scale environmental context` == "human-associated habitat")) ~ "over 3",
    str_detect(sample_id, "SAMN04436") ~ "under 3",
    (str_detect(sample_id, "SAMN04571") & (`broad-scale environmental context` == "human gut")) ~ "over 3",
    str_detect(sample_id, "SAMN04978193") ~ "over 3",
    str_detect(sample_id, "SAMN0506") ~ "over 3",
    str_detect(sample_id, "SAMN05194") ~ "over 3",
    str_detect(sample_id, "SAMN05214") ~ "over 3",
    str_detect(sample_id, "SAMN053014|SAMN053015") ~ "under 3",
    str_detect(sample_id, "SAMN0535") ~ "over 3",
    str_detect(sample_id, "SAMN055785") ~ "over 3",
    str_detect(sample_id, "SAMN05780991") ~ "under 3",
    (str_detect(sample_id, "SAMN058") & (`broad-scale environmental context` == "ENVO:2100002")) ~ "under 3",
    (str_detect(sample_id, "SAMN060") & (`local-scale environmental context` == "ankylosing spondylitis patient")) ~ "over 3",
    (str_detect(sample_id, "SAMN060") & (`geographic location` == "USA: Chicago")) ~ "over 3",
    (str_detect(sample_id, "SAMN061") & (`local-scale environmental context` == "ankylosing spondylitis patient")) ~ "over 3",
    str_detect(sample_id, "SAMN061153|SAMN061154|SAMN061155") ~ "over 3",
      (str_detect(sample_id, "SAMN061") & (`nucleic_acid` == "dna")) ~ "over 3",
    str_detect(sample_id, "SAMN07501") ~ "under 3",
    str_detect(sample_id, "SAMN079777") ~ "under 3",
    str_detect(sample_id, "SAMN0824") ~ "over 3",
    str_detect(sample_id, "SAMN08533") ~ "over 3",
    str_detect(sample_id, "SAMN09062") ~ "over 3",
    str_detect(sample_id, "SAMN09711") ~ "over 3",
    str_detect(sample_id, "SAMN09866") ~ "over 3",
    str_detect(sample_id, "SAMN10720") ~ "over 3",
    str_detect(sample_id, "SAMN1100") ~ "over 3",
    str_detect(sample_id, "SAMN1238") ~ "over 3",
    str_detect(sample_id, "SAMN1465") ~ "over 3",
    str_detect(sample_id, "SAMD000361|SAMD000362|SAMD000363|SAMD000364") ~ "over 3",
    str_detect(sample_id, "SAMD00036524") ~ "over 3",
    str_detect(sample_id, "SAMD00036525") ~ "over 3",
    str_detect(sample_id, "SAMD00036526") ~ "under 3",
    str_detect(sample_id, "SAMD00036527") ~ "over 3",
    str_detect(sample_id, "SAMD00036528") ~ "over 3",
    str_detect(sample_id, "SAMD00036529") ~ "under 3", # exactly 3
    str_detect(sample_id, "SAMD00036530") ~ "under 3",
    str_detect(sample_id, "SAMD00036531") ~ "over 3",
    str_detect(sample_id, "SAMD00036532") ~ "under 3",
    str_detect(sample_id, "SAMD00036533") ~ "over 3",
    str_detect(sample_id, "SAMD00036534") ~ "under 3",
    str_detect(sample_id, "SAMD00036535") ~ "under 3",
    str_detect(sample_id, "SAMD00036536") ~ "over 3",
    TRUE ~ "NA"
)) %>%
    relocate(under_over_3, .after = `age_years_nayfach`) %>%
    relocate(country, .after = `geographic location`)

```

```{r}
srb_lite_notes_ages_countries <- srb_lite_notes_ages %>%
  mutate(country = case_when(
    str_detect(sample_id, "SAMD00036") ~ "Japan",
    str_detect(sample_id, "SAMN00265") ~ "Finland",
    str_detect(sample_id, "SAMN00715|SAMN00791|SAMN00792|SAMN00993") ~ "China",
    str_detect(sample_id, "SAMN0295426") ~ "United States of America", # IRB approval at Stanford Uni, "one subject had traveled to Brazil and France", suggesting they are from the U.S.A.
    str_detect(sample_id, "SAMN0377735") ~ "Italy", # pilot for larger study with IRB approval at Uni of Trento
    str_detect(sample_id, "SAMN04978193") ~ "United Kingdom", # samples from UK adult twin registry
    str_detect(sample_id, "SAMN0824") ~ "United States of America",
    str_detect(sample_id, "SAMN08533") ~ "United States of America", # recruited at Stanford Uni
    str_detect(sample_id, "SAMN10720|SAMN1238") ~ "United States of America", # recruited at Stanford Uni
    str_detect(sample_id, "SAMN03279760|SAMN03281006|SAMN03281007|SAMN03281010|SAMN03281016|SAMN03281018|SAMN03281020|SAMN03281021") ~ "Malaysia",
    str_detect(sample_id, "SAMN03281009|SAMN03281011") ~ "India",
    str_detect(sample_id, "SAMN03281017|SAMN03281019") ~ "China",
    TRUE ~ country))

#write.csv(srb_lite_notes_ages_countries, file = "srb_lite_notes_ages_countries_04-10-2026.csv", row.names = FALSE)
saveRDS(srb_lite_notes_ages_countries, file = paste0(dataDir, "/srb_lite_notes_ages_countries.rds"))


```

## add subject info
#ie is it longitudinal? 
```{r}

```

# make .txt file of the biosample IDs
```{r}
IDs <- srb_lite_notes_ages_countries$sample_id
write(IDs, file = "biosample_IDs_04-13-2026.txt")
```



# making sure that the query IDs are unique since that will be used to filter the compiled csv files for only the samples we want to keep
# code generated with Claude AI
```{r}
# extract Query IDs from their list and turn df into long format
df_long <- srb_lite_notes_ages_countries %>%
  mutate(query_ids = str_extract_all(query_ids, "[A-Z0-9]+")) %>%  # parse the string lists
  unnest(query_ids)

# are any of these IDs *not* unique?
duplicated_ids <- df_long %>%
  count(query_ids) %>%
  filter(n > 1)

## second way of checking, splitting by commas in the list rather than regex
all_ids <- unlist(strsplit(srb_lite_notes_ages_countries$query_ids, ","))
all_ids <- trimws(all_ids)  # remove whitespace
all_ids <- gsub("\\[", "", all_ids)
all_ids <- gsub("\\]", "", all_ids)
all_ids <- gsub("'", "", all_ids)  # remove brackets and quotes

unique_ids <- unique(all_ids)
duplicated_ids <- all_ids[duplicated(all_ids)]

# make sure these methods give the same results
ids_regex <- df_long %>%
  pull(query_ids) %>%
  sort()

# method 2 - comma split
ids_split <- all_ids %>% sort()

# compare
identical(ids_regex, ids_split)
```


## duplicate AS studies; both have 97 samples; which is the one that has assemblies? kept the one with hits in a previous iteration of the pipeline (all 72k samples)
```{r}
AS_samples <- srb_lite %>%
  filter((str_detect(sample_id, "SAMN060|SAMN0634") & (`local-scale environmental context` == "ankylosing spondylitis patient")))

hits <- read_csv(paste0(dataDir, "closestRefInfo_allScoreThresholdHits_wSingleClosestRef_no_2018.csv"))

AS_hits <- hits %>%
  filter(biosample_id %in% AS_samples$sample_id)

all_hits <- hits %>%
  filter(biosample_id %in% srb_lite_notes_ages_countries$sample_id)
```

# trying to see how many biosample_ids are associated with multiple query ids
# does having multiple query IDs mean there are duplicates in the dataset? -> you would see multiple of the same species/hits from the same sample but different assembly/sequencing run
# code generated with help from Claude
```{r}
mult_queries <- srb_lite_notes_ages_countries$sample_id[sapply(strsplit(srb_lite_notes_ages_countries$query_ids, ","), length) > 1]

#extract the sample_id and query_ids for samples with multiple queries
samples_w_mult_queries <- srb_lite_notes_ages_countries %>% filter(sample_id %in% mult_queries) %>% select(sample_id, query_ids)

queries_long <- samples_w_mult_queries %>%
  mutate(query_ids = str_extract_all(query_ids, "[A-Z0-9]+")) %>%  # parse the string lists
  unnest(query_ids)

hits_for_assemblies <- read_csv(paste0(dataDir, "closestRefInfo_allScoreThresholdHits_no_2018_assemblies.csv"))

# are there samples that have hits from more than one different query?
samples_w_multiple_query_hits <- hits_for_assemblies %>% filter(query_id %in% queries_long$query_ids) %>%
  rename(query_ids = query_id) %>%
  left_join(queries_long, by = "query_ids") %>%
  # don't need to worry about the ones hitting moorella thermoacetica
  filter(distance < 2.5 | is.na(distance)) %>%
  group_by(sample_id) %>%
  summarise(n_query_ids = n_distinct(query_ids)) %>%
  filter(n_query_ids > 1)

#now, find which hits these are -> which samples with multiple query IDs actually had both hit something
just_filtered <- hits_for_assemblies %>% filter(query_id %in% queries_long$query_ids) %>%
  rename(query_ids = query_id) %>%
  left_join(queries_long, by = "query_ids") %>%
  filter(sample_id %in% samples_w_multiple_query_hits$sample_id) %>%
  group_by(sample_id) %>%
  summarise(query_list = list(unique(query_ids)))

# make df for .txt file of these queries that can be used with snakemake rule
qoi <- just_filtered %>%
  mutate(query_list = str_extract_all(query_list, "[A-Z0-9]+")) %>%
  unnest(query_list)

# queries that both have hits and are from one sample
write(qoi$query_list, file = "queries_dups_04-15-2026.txt")

# there are 59 biosamples that have several queries that both have hits
# all of these look like viable assemblies aside from 2 samples. 
# instead select by the query with more hits ??
which_query <- hits_for_assemblies %>%
  filter(query_id %in% qoi$query_list)

## instead of trying to choose them below, I can do this after removing other duplicates... left_join the sample ID then group_by
# SAMD00036196 OQGT01 (the other has 4 genes on one contig)
# SAMD00036214 OQHG01 (higher bit score for the hit to the same species)
# SAMD00036228 OPYT01 (the other has 3 genes on one contig)
# SAMD00036229 OQCB01 (shorter distance to the same species)
# SAMD00036251 OQAD01 (had a slightly higher bit score for a hit to the same species)
# SAMD00036316 OQEN01 (the other only had one species aside from moorella with a bit score barely over 100)
# SAMD00036330 OQDZ01 (higher bit score for same species)
# SAMD00036337 OQDT01 (3 hits to one contig for bilophila species as opposed to one)
# SAMD00036377 OQNO01 (much longer length hits)
# SAMD00036381 OQLH01 (much longer length hits)
# SAMD00036386 OQNI01 (subunits are mapping to same species)
# SAMD00036430 OQLI01 (more non-sdup hits)
# SAMD00036432 OQNV01 (more species)
# SAMEA1963786 OPRH01 (longer hits)



```


