---
title: "tree_visualization"
output: html_document
date: "2024-06-21"
---

```{r}
library(ggplot2)
library(treeio)
library(ggtree)
library(dplyr)
library(ape)
library(phytools)
library(readr)
library(seqinr)
library(xlsx)
library(readxl)
```

# dfs to load
```{r}
rooted_ref_tree_f = read.tree("../../workflow/out/raxmlOutput/RAxML_labelledTree_noBootstrap_capsule_plus_wgs.newick")

rooted_result_tree_f = read.tree("../../workflow/out/raxmlOutput/RAxML_labelledTree_noBootstrap_capsule_plus_wgs.newick")

closestRefInfo_allScoreThresholdHits_wSingleClosestRef <- read_csv("../workflow/out/treeInfo/closestRefInfo_allScoreThresholdHits_wSingleClosestRef_capsule_plus_wgs.csv")

otu_hits_table <- read_csv("hits_per_sequence.csv", 
    col_types = cols(...1 = col_skip()))

sample_hits_table <- read_csv("sample_hits_table.csv", 
    col_types = cols(...1 = col_skip()))

tax_df_closest_hits <- read_excel("taxonomy_df_of_closest_hit_sequences.xlsx", 
                                  sheet = "Sheet2", col_types = c("skip", 
                                                                  "text", "text", "text", "text", "text", 
                                                                  "text", "text", "text", "text", "text", 
                                                                  "text", "text"))

```


# otu_hits_table: number of hit sequences (including duplicates) associated with each reference sequence (by shortest branch distance) for all samples
# sample_hits_table: number of samples ("Biosample" as defined by NCBI) containing each hit sequence (rounded to shortest reference sequence)
```{r}
otu_hits_table <- as.data.frame(table(subset(closestRefInfo_allScoreThresholdHits_wSingleClosestRef$single_closest_ref, closestRefInfo_allScoreThresholdHits_wSingleClosestRef$distance < 2.5 | is.na(closestRefInfo_allScoreThresholdHits_wSingleClosestRef$distance))))
otu_hits_table <- rename(otu_hits_table, tip_label = Var1)

# write.csv(otu_hits_table, "hits_per_sequence.csv")

sample_hits_table <- closestRefInfo_allScoreThresholdHits_wSingleClosestRef %>%
  filter(distance < 2.5 | is.na(distance)) %>%
  group_by(single_closest_ref) %>%
  summarize(n_samples = n_distinct(biosample_id))

sample_hits_table <- rename(sample_hits_table, tip_label = single_closest_ref)

# write.csv(sample_hits_table, "sample_hits_table.csv")
```

# Make a new taxonomy df for each sequence in the ref tree, including Anantharaman 2018 seq.
```{r}
# read in the fasta file for ref seqs, complete with taxonomy descriptions from Mueller 2015 paper

# hack way for reading in a fasta file
ref_fa <- read_delim("Muller2015_reference_seqs_with_taxonomy.txt", delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)

ref_fa <- na.omit(ref_fa) #lol it worked!
ref_fa <- rename(ref_fa, seq_abbreviation = X1,
                 seq_name = X2,
                 strain_name = X3,
                 environment = X4,
                 description = X5)

id <- as.character(ref_fa[1, "description"])
strsplit(id, ";")

get_taxa <- function(description, level) {
  taxa <- strsplit(as.character(description), ";")[[1]][level]
  return(taxa)
}

ref_fa <- ref_fa %>%
  rowwise() %>%
  mutate(tax_1 = get_taxa(description, 1),
         tax_2 = get_taxa(description, 2),
         tax_3 = get_taxa(description, 3),
         tax_4 = get_taxa(description, 4),
         tax_5 = get_taxa(description, 5))
# WEEEE

# ok i can bind the otu_hits_table to ref_fa using the seq_abbreviation column in ref_fa by creating a new column that matches this format in otu_hits_table

make_seq_abbreviation <- function(label) {
  label <- strsplit(as.character(label), "_")[[1]]
  abbreviation <- tail(label, n=1)
  abbreviation <- paste(">", abbreviation, sep="")
  
  return(abbreviation)
}

ref_seqs <- data_frame(tip_label = rooted_ref_tree_f$tip.label)
all_seqs <- data_frame(tip_label = rooted_result_tree_f$tip.label)

ref_seqs <- ref_seqs %>%
  rowwise() %>%
  mutate(seq_abbreviation = make_seq_abbreviation(tip_label))

# manually correct seq_abbreviation for Candidatus_Desulfovibrio_trichonymphae
ref_seqs$seq_abbreviation[ref_seqs$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$seq_abbreviation[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]
ref_tax <- merge(ref_fa, ref_seqs, by = "seq_abbreviation")

anantharaman_seqs <- read_delim("../config/Anantharaman2018_dsrA_dsrB.faa", delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
anantharaman_seqs <- subset(anantharaman_seqs, grepl(">", X1))
anantharaman_seqs <- rename(anantharaman_seqs, seq_abbreviation = X1)

make_tip_label <- function(seq_name){
  label <- gsub(">", "", as.character(seq_name))
  label <- paste("QUERY___", label, sep = "")
  return(label)
}

anantharaman_seqs <- anantharaman_seqs %>%
  rowwise() %>%
  mutate(tip_label = make_tip_label(seq_abbreviation))

# bind taxa data to df with tip labels
tax_df <- merge(anantharaman_seqs, ref_tax, by = c("tip_label", "seq_abbreviation"), all = T)

tax_df[is.na(tax_df)] <- " Unclassified"

anantharaman_seqs_in_tree <- subset(anantharaman_seqs, tip_label %in% all_seqs$tip_label)

# write.csv(tax_df, "dsrAB_taxonomy_table.csv")
# write.csv(ref_seqs, "reference_sequence_naming_table.csv")
# write.csv(anantharaman_seqs_in_tree, "anantharaman_seqs_in_tree.csv")

```

# Manually update taxonomy info for named Anantharaman seqs in tax_df
```{r}
tax_df$tax_1[tax_df$tip_label == "QUERY___Bilophila_wadsworthia_dsrA"] <- tax_df$tax_1[tax_df$tip_label == "Bilophila_wadsworthia__BilWads6"]
tax_df$tax_2[tax_df$tip_label == "QUERY___Bilophila_wadsworthia_dsrA"] <- tax_df$tax_2[tax_df$tip_label == "Bilophila_wadsworthia__BilWads6"]
tax_df$tax_3[tax_df$tip_label == "QUERY___Bilophila_wadsworthia_dsrA"] <- tax_df$tax_3[tax_df$tip_label == "Bilophila_wadsworthia__BilWads6"]
tax_df$tax_4[tax_df$tip_label == "QUERY___Bilophila_wadsworthia_dsrA"] <- tax_df$tax_4[tax_df$tip_label == "Bilophila_wadsworthia__BilWads6"]
tax_df$tax_5[tax_df$tip_label == "QUERY___Bilophila_wadsworthia_dsrA"] <- tax_df$tax_5[tax_df$tip_label == "Bilophila_wadsworthia__BilWads6"]

tax_df$tax_1[tax_df$tip_label == "QUERY___Bilophila_wadsworthia_dsrB"] <- tax_df$tax_1[tax_df$tip_label == "Bilophila_wadsworthia__BilWads6"]
tax_df$tax_2[tax_df$tip_label == "QUERY___Bilophila_wadsworthia_dsrB"] <- tax_df$tax_2[tax_df$tip_label == "Bilophila_wadsworthia__BilWads6"]
tax_df$tax_3[tax_df$tip_label == "QUERY___Bilophila_wadsworthia_dsrB"] <- tax_df$tax_3[tax_df$tip_label == "Bilophila_wadsworthia__BilWads6"]
tax_df$tax_4[tax_df$tip_label == "QUERY___Bilophila_wadsworthia_dsrB"] <- tax_df$tax_4[tax_df$tip_label == "Bilophila_wadsworthia__BilWads6"]
tax_df$tax_5[tax_df$tip_label == "QUERY___Bilophila_wadsworthia_dsrB"] <- tax_df$tax_5[tax_df$tip_label == "Bilophila_wadsworthia__BilWads6"]

tax_df$tax_1[tax_df$tip_label == "QUERY___Desulfosporosinus_orientis_dsrA"] <- tax_df$tax_1[tax_df$tip_label == "Desulfosporosinus_orientis__DesOrie4"]
tax_df$tax_2[tax_df$tip_label == "QUERY___Desulfosporosinus_orientis_dsrA"] <- tax_df$tax_2[tax_df$tip_label == "Desulfosporosinus_orientis__DesOrie4"]
tax_df$tax_3[tax_df$tip_label == "QUERY___Desulfosporosinus_orientis_dsrA"] <- tax_df$tax_3[tax_df$tip_label == "Desulfosporosinus_orientis__DesOrie4"]
tax_df$tax_4[tax_df$tip_label == "QUERY___Desulfosporosinus_orientis_dsrA"] <- tax_df$tax_4[tax_df$tip_label == "Desulfosporosinus_orientis__DesOrie4"]
tax_df$tax_5[tax_df$tip_label == "QUERY___Desulfosporosinus_orientis_dsrA"] <- tax_df$tax_5[tax_df$tip_label == "Desulfosporosinus_orientis__DesOrie4"]

tax_df$tax_1[tax_df$tip_label == "QUERY___Archaeoglobus_sulfaticallidus_dsrA"] <- tax_df$tax_1[tax_df$tip_label == "Archaeoglobus_sulfaticallidus__ArgSulfa"]
tax_df$tax_2[tax_df$tip_label == "QUERY___Archaeoglobus_sulfaticallidus_dsrA"] <- tax_df$tax_2[tax_df$tip_label == "Archaeoglobus_sulfaticallidus__ArgSulfa"]
tax_df$tax_3[tax_df$tip_label == "QUERY___Archaeoglobus_sulfaticallidus_dsrA"] <- tax_df$tax_3[tax_df$tip_label == "Archaeoglobus_sulfaticallidus__ArgSulfa"]
tax_df$tax_4[tax_df$tip_label == "QUERY___Archaeoglobus_sulfaticallidus_dsrA"] <- tax_df$tax_4[tax_df$tip_label == "Archaeoglobus_sulfaticallidus__ArgSulfa"]
tax_df$tax_5[tax_df$tip_label == "QUERY___Archaeoglobus_sulfaticallidus_dsrA"] <- tax_df$tax_5[tax_df$tip_label == "Archaeoglobus_sulfaticallidus__ArgSulfa"]

tax_df$tax_1[tax_df$tip_label == "QUERY___Thermodesulfobium_narugense_dsrA"] <- tax_df$tax_1[tax_df$tip_label == "Thermodesulfobium_narugense__TrfNaru2"]
tax_df$tax_2[tax_df$tip_label == "QUERY___Thermodesulfobium_narugense_dsrA"] <- tax_df$tax_2[tax_df$tip_label == "Thermodesulfobium_narugense__TrfNaru2"]
tax_df$tax_3[tax_df$tip_label == "QUERY___Thermodesulfobium_narugense_dsrA"] <- tax_df$tax_3[tax_df$tip_label == "Thermodesulfobium_narugense__TrfNaru2"]
tax_df$tax_4[tax_df$tip_label == "QUERY___Thermodesulfobium_narugense_dsrA"] <- tax_df$tax_4[tax_df$tip_label == "Thermodesulfobium_narugense__TrfNaru2"]
tax_df$tax_5[tax_df$tip_label == "QUERY___Thermodesulfobium_narugense_dsrA"] <- tax_df$tax_5[tax_df$tip_label == "Thermodesulfobium_narugense__TrfNaru2"]

tax_df$seq_abbreviation[tax_df$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$seq_abbreviation[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]
tax_df$seq_name[tax_df$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$seq_name[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]
tax_df$strain_name[tax_df$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$strain_name[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]
tax_df$environment[tax_df$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$environment[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]
tax_df$description[tax_df$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$description[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]
tax_df$tax_1[tax_df$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$tax_1[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]
tax_df$tax_2[tax_df$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$tax_2[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]
tax_df$tax_3[tax_df$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$tax_3[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]
tax_df$tax_4[tax_df$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$tax_4[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]
tax_df$tax_5[tax_df$tip_label == "Candidatus_Desulfovibrio_trichonymphae"] <- ref_fa$tax_5[ref_fa$seq_name == "Candidatus Desulfovibrio trichonymphae"]

```

# Trim results tree to only include reference sequences. Save data which will be needed to create the paneled tree plot. 
```{r}
ref_tips <- append(ref_seqs$tip_label, anantharaman_seqs_in_tree$tip_label)
query_tips <- setdiff(all_seqs$tip_label, ref_tips)

# function to auto generate list of groups for different taxon levels courtsey of the internet woohoo
taxa_groups_df <- subset(tax_df, tax_5 != "" & !is.na(tax_5) & !is.na(tip_label))

write.csv(taxa_groups_df, "taxa_groups_df.csv")

# closest hits (reference sequences) only tree
not_closest_hits <- setdiff(rooted_result_tree_f$tip.label, otu_hits_table$tip_label)
closest_hits_tree <- drop.tip(rooted_result_tree_f, not_closest_hits)

write.tree(closest_hits_tree, "closest_hits_tree.newick")

tax_df_closest_hits <- subset(tax_df, tip_label %in% closest_hits_tree$tip.label)

tax_df_closest_hits <- tax_df_closest_hits %>%
  mutate(source = ifelse(grepl("QUERY", tip_label), 
                         "Anantharaman 2018", 
                         "Müller 2015"))

write.csv(tax_df_closest_hits, "tax_df_closest_hits.csv")

```

