---
title: "SSRMprevalence"
output: html_document
author: "Elizabeth Daigle"
date: "16 April 2026"
---
# this .rmd gets the sample prevalence of SSRMs by age 
# also calculates number of countries represented 
# also gets numbers for results section

# Read in packages and functions
```{r}
source(paste0(here::here(), "/1-Config/config.R"))
```
# Read in metadata and hits from dm_NCBI.Rmd
```{r}
metadata_final <- readRDS(paste0(dataDir, "/metadata_final.rds"))

# remove hits that were not inserted properly or are too far (Moorella thermoacetica)
hits_final_ages <- readRDS(paste0(dataDir, "/hits_final_ages_no_MA.rds")) %>%
  # removes one of archaeoglobus hits that wasn't removed in other spots but had no bearing on results
  filter(is.na(distance) | distance <= 2.50) %>%
  #remove one of the hits from the three exact duplicates ("ORJZ01.17253.1_1/2-182", "OWUN01.41726.1_1/1-246", "OWAH01.53685.1_1/1-179")
  distinct()

## 4093 is the number of "putative dsrAB hits" -> filtered for no closest reference, distance >2.5 but still contains dups (sdups and edups for the sequence itself)
```

# calculate SSRM sample prevalence
# grouping by biosample avoids the need to remove the excess query IDs (some samples have multiple)
```{r}
n_samples <- n_distinct(metadata_final$sample_id)

n_samples_under_3 <- n_distinct(metadata_final$sample_id[metadata_final$under_over_3 == "under 3"])

n_samples_over_3 <- n_distinct(metadata_final$sample_id[metadata_final$under_over_3 == "over 3"])

overall <- hits_final_ages %>%
  summarize(n_biosamples_hits = n_distinct(biosample_id)) %>%
  mutate(n_biosamples_total = n_samples) %>%
  mutate(sample_SSRM_prevalence = round(n_biosamples_hits/n_biosamples_total, digits = 4)) %>%
  mutate(percentage = round(sample_SSRM_prevalence*100, digits = 2)) %>%
  mutate(group = "All ages")

under_3 <- hits_final_ages %>%
  filter(under_over_3 == "under 3") %>%
  summarize(n_biosamples_hits = n_distinct(biosample_id)) %>%
  mutate(n_biosamples_total = n_samples_under_3) %>%
  mutate(sample_SSRM_prevalence = round(n_biosamples_hits/n_biosamples_total, digits = 4)) %>%
  mutate(percentage = round(sample_SSRM_prevalence*100, digits = 2)) %>%
  mutate(group = "Subjects under age 3")

over_3 <- hits_final_ages %>%
  filter(under_over_3 == "over 3") %>%
  summarize(n_biosamples_hits = n_distinct(biosample_id)) %>%
  mutate(n_biosamples_total = n_samples_over_3) %>%
  mutate(sample_SSRM_prevalence = round(n_biosamples_hits/n_biosamples_total, digits = 4)) %>%
  mutate(percentage = round(sample_SSRM_prevalence*100, digits = 2)) %>%
  mutate(group = "Subjects over age 3")

prevalence <- bind_rows(overall, under_3, over_3) %>%
  relocate(group, .before = n_biosamples_hits)
```

# what studies are the infant hits from 
```{r}
under_3_study <- hits_final_ages %>%
  filter(under_over_3 == "under 3") %>%
  group_by(biosample_id) %>%
  summarize(n_biosamples_hits = n()) %>%
  left_join(metadata_final, by = c("biosample_id"="sample_id"))
```

# make plot of under 3 age distribution
```{r}
under_three_distribution <- data.frame(
  age_bin = c("0-1", "1-2", "2-3"),
  n_total = c(1109,377,96)
)

age_color <- c("#B4D6E3", "#7F9DB1", "#0D3B66")

hits_per_bin <- hits_final_ages %>%
  filter(under_over_3 == "under 3") %>%
  group_by(biosample_id) %>%
  summarize(n_hits = n())

#SAMN04161035 is 0-1
hits_per_bin_w_age <- hits_per_bin %>%
  left_join((metadata_final %>% select(c(sample_id,`host age`))), by = c("biosample_id"="sample_id")) %>%
  mutate(age_bin = case_when(
    is.na(`host age`) ~ "0-1",
    as.numeric(`host age`) <= 364 ~ "0-1",
    as.numeric(`host age`) <= 729 ~ "1-2",
    #generally close enough to being three...
    as.numeric(`host age`) >= 730 ~ "2-3")) %>%
  group_by(age_bin) %>%
  summarize(n_positive=n()) %>%
  left_join(under_three_distribution, by = "age_bin") %>%
  mutate(proportion_pos = n_positive/n_total)

total <- ggplot(hits_per_bin_w_age, aes(x = age_bin, y = n_total, fill = age_bin)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = age_color) +
  labs(
    x = "Age",
    y = "Number of samples"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    plot.title = element_blank()
  )

prop_pos <- ggplot(hits_per_bin_w_age, aes(x = age_bin, y = proportion_pos*100, fill = age_bin)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = age_color) +
  labs(
    x = "Age",
    y = "SSRM positive samples (%)"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    plot.title = element_blank()
  )

library(patchwork)

plot <- total+prop_pos

ggsave(plot = plot, width = 8, height = 5, file = paste0(figureDir, "under3_barplot.jpeg"))
```


# number of countries 
```{r}
countries <- metadata_final %>% 
  mutate(country = case_when(
    str_detect(country, "USA") ~ "United States of America",
    TRUE ~ country)) %>%
  group_by(country) %>% 
  summarize(n=n())
```


# unique hits
```{r}
# group the hits by biosample 
# select one domain and one subunit 
unique_hits <- hits_final_ages %>%
  mutate(contig = str_extract(hit_id, "(?<=\\.).*(?=_)")) %>%
  group_by(contig, biosample_id) %>%
  summarize(n=n())
#2033 hits across all of the samples

# but how many that are not sdups or edups? 
noDups <- read_csv(paste0(treeInfoDir, "queryDistanceInfo_no_2018_assemblies.csv")) 
#1481 unique dsrAB sequences
```
# for results, desulfovibrionaceae family and firmicutes group prevalences
```{r}
firmicute <- c("Anaerobic_bacterium_EtOH8__Bv2EtOH2", "Desulfitibacter_alkalitolerans__DfiAlka2", "Gordonibacter_pamelaeae__GrdPame5", "Gordonibacter_pamelaeae_7-10-1-b__FP929047", "Thermodesulfobium_narugense__CP002690", "MSW_digester_clone__Unc01cxc", "Anaerobic_bacterium_sk.prop8__Bv2Prop2", "Bioreactor_clone__UncS2029", "Arctic_sediment_clone_AH5_5__AtcSed67", "Desulfurispora_thermophila__DprTher2", "Desulfotomaculum_hydrothermale__DscHydr2", "Polluted_aquifer_clone_LGWK16__UncS1633", "Desulfosporosinus_youngiae__DfpYoun5")

desulfo <- c("Bilophila_sp._4_1_30__BilSpec3", "Bilophila_wadsworthia__BilWads6", "Desulfovibrio_piger__DslPige2", "Desulfovibrio_piger__DslPige3", "Desulfovibrio_piger__DslPige4", "Desulfovibrio_desulfuricans__DslDes17", "Desulfovibrio_desulfuricans_subsp._desulfuricans_str._ATCC_27774__CP001358", "Desulfovibrio_simplex__DslSimpl", "Desulfovibrio_africanus__DslAfri3", "uncultured_bacterium__EF645665", "uncultured_bacterium__EF645667", "uncultured_bacterium__EF645664", "Desulfovibrio_sp._6_1_46AFAA__DsvSp231", "Desulfovibrio_sp._3_1_syn3__DsvSp230", "Desulfovibrio_sp._Dsv1__DsvSp233", "Desulfovibrio_vulgaris__DQ826729", "Desulfovibrio_desulfuricans__DQ092635", "Candidatus_Desulfovibrio_trichonymphae", "Desulfovibrio_desulfuricans__DslDes16_DsvAlas8", "Desulfovibrio_carbinoliphilus__JQ031012")

# making sure I didn't forget any of the 33
# hits_d_f <- hits_final_ages %>% 
#   filter(single_closest_ref %in% c(firmicute,desulfo))
# there are 33 unique single closest refs which is what I expect

hits_firmicute <- hits_final_ages %>%
  filter(single_closest_ref %in% firmicute)
(prev_firmicute <- n_distinct(hits_firmicute)/n_distinct(hits_final_ages))

hits_desulfo <- hits_final_ages %>%
  filter(single_closest_ref %in% desulfo)
(prev_desulfo <- n_distinct(hits_desulfo)/n_distinct(hits_final_ages))

(prev_firmicute + prev_desulfo)

```

# looking at age spread under 3 years
```{r}
diabimmune <- metadata_final %>% 
  filter(str_detect(notes, "DIABIMMUNE")) %>%
  mutate(age_group = case_when(
    as.numeric(`host age`) <= 364 ~ "0-1",
    as.numeric(`host age`) <= 729 ~ "1-2",
    as.numeric(`host age`) >= 730 ~ "2+")) %>%
  count(age_group)

#from Backhed et al. 2015 supp. table 1; child ages at 12m sampling time point
infantgut <- as.numeric(c(380,363,363,369,366,409,359,369,369,376,364,364,374,367,375,364,366,372,364,366,
               371,359,369,374,378,365,367,378,354,366,369,361,357,360,396,381,377,371,365,358,
               362,385,363,372,371,367,373,366,385,364,364,364,376,372,380,372,370,371,365,361,
               360,364,364,358,360,367,367,371,355,364,372,352,372,353,363,365,365,357,358,359,
               358,369,351,380,362,376,367,368,366,364,361,368,360,365,367,398,371,359))
sum(infantgut>=365)

```


