---
title: "abundance_figures"
date: "25 November 2025"
output: html_document
author: "Elizabeth Daigle"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Read in packages and functions
```{r}
source(paste0(here::here(), "/0-Config_Functions/0-Config.R"))
```

#here making new plots
Making viridis color palette for plots
```{r}
# Generate a viridis palette with 20 colors
viridis_palette <- viridis(20)
viridis_palette

boxplot_colors <- c(viridis_palette[6], viridis_palette[16])

jitter_colors <- c("#4477AA", "#59A075", "#AA3377", "#DDAA33", "#6F7042")
```

# Plot the number of unique genes for each subject 
```{r}
plot_count_dsr <- samples_unique_genes_with_names %>%
# find number of unique genes for each sample
  group_by(subject, sample, Type) %>%
  summarize(n=n(), .groups="drop") 

plot_count_dsr_median <- samples_unique_genes_with_names %>%
# find number of unique genes for each sample
  group_by(subject, sample, Type) %>%
  summarize(n=n(), .groups="drop") %>%
# find median of each type
  group_by(subject, Type) %>% 
  summarize(median_n = median(n))

plot_count_dsr_median %>%
  ggplot(aes(x = Type, y = median_n)) +
  geom_boxplot() + 
  geom_jitter(aes(color = Type),
              alpha = 0.7, 
              size = 3,
              position = position_jitter(width = 0.15, height = 0)) +
  scale_color_manual(values = jitter_colors) +
  labs(
    title = "Median dsrAB gene count by sample type",
    x = "Sample Type",
    y = "Subject Median Gene Count",
    color = "Type"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 24, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 24),
    axis.title.x = element_text(size = 24),
    axis.title.y = element_text(size = 24),
    legend.text = element_text(size = 24),
    legend.title = element_text(size = 24),
    strip.placement = "outside",
    strip.background = element_blank(),
    plot.title = element_text(size = 28, hjust = 0.5),
    plot.subtitle = element_text(size = 24, hjust = 0.5))

plot_count_dsr %>%
  ggplot(aes(x = Type, y = n)) +
  geom_boxplot() + 
  geom_jitter(aes(color = Type),
              alpha = 0.7, 
              size = 3,
              position = position_jitter(width = 0.15, height = 0)) +
  scale_color_manual(values = jitter_colors) +
  facet_wrap(~ subject,
              #scales = "free_y",
              strip.position = "bottom",
              labeller = labeller(subject = function(x) paste("Subject", x))) +
  labs(
    title = "Unique dsrAB gene count by sample type across subjects",
    x = "Sample Type",
    y = "Gene Count",
    color = "Type"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 24, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 24),
    axis.title.x = element_text(size = 24),
    axis.title.y = element_text(size = 24),
    legend.text = element_text(size = 24),
    legend.title = element_text(size = 24),
    strip.placement = "outside",
    strip.background = element_blank(),
    plot.title = element_text(size = 28, hjust = 0.5),
    plot.subtitle = element_text(size = 24, hjust = 0.5))
```
# Change species labels
```{r}
species_labels <- c("QUERY___Bilophila_wadsworthia_dsrA" = "Bilophila wadsworthia dsrA",
                    "Desulfitibacter_alkalitolerans__DfiAlka2" = "Desulfitibacter alkalitolerans",
                    "Desulfovibrio_piger__DslPige2" = "Desulfovibrio piger",
                    "Bilophila_wadsworthia__BilWads6" = "Bilophila wadsworthia 6",
                    "uncultured_bacterium__EF645667" = "Desulfovibrio desulfuricans",
                    "Bilophila_sp._4_1_30__BilSpec3" = "Bilophila species 3",
                    "Gordonibacter_pamelaeae__GrdPame5" = "Gordonibacter pamelaeae",
                    "QUERY___Moorella_thermoacetica_2_dsrA" = "Moorella thermoacetica")
```

# plot RPKM for every hit to compare between species; facet by location and subject
```{r}
samples_unique_genes_with_names %>%
  filter(!single_closest_ref %in% c("QUERY___Ley3_66761_scaffold_1025_dsrA",
                       "QUERY___Ley3_66761_scaffold_1830_dsrB",
                       #"Gordonibacter_pamelaeae__GrdPame5",
                       #"QUERY___Moorella_thermoacetica_2_dsrA",
                       "Anaerobic_bacterium_sk.prop8__Bv2Prop2")) %>%
  ggplot(aes(x = single_closest_ref, y = log10(RPKM), fill = Type)) +
  geom_boxplot(color = "black",
               position = position_dodge(width = 0.8), size = 0.8) +
  geom_jitter(aes(color = Type),
            alpha = 0.9, size = 2,
            position = position_jitterdodge(
        jitter.width = 0.5,
        dodge.width = 0.8)) +
  scale_color_manual(values = jitter_colors) +
  #get rid of box fill
  scale_fill_manual(values = c("transparent", "transparent", "transparent", "transparent", "transparent"), guide = "none") +
  scale_x_discrete(labels = species_labels) + 
  guides(fill = "none") +
  facet_grid(subject ~ Type, scale = "free_y") +
  labs(
    title = "RPKM for unique dsrAB genes by species across subjects and sample locations",
    x = "Closest Reference",
    y = "log10(RPKM)",
    color = "Sample Location"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 60, hjust = 1),
    plot.title = element_text(size = 20, hjust = 0.5),
    plot.subtitle = element_text(size = 16, hjust = 0.5)
  )
```
# plot RPKM for every hit to compare between location; facet by species and subject
```{r}
samples_unique_genes_with_names %>%
  filter(!single_closest_ref %in% c("QUERY___Ley3_66761_scaffold_1025_dsrA",
                       "QUERY___Ley3_66761_scaffold_1830_dsrB",
                       #"Gordonibacter_pamelaeae__GrdPame5",
                       #"QUERY___Moorella_thermoacetica_2_dsrA",
                       "Anaerobic_bacterium_sk.prop8__Bv2Prop2")) %>%
  ggplot(aes(x = Type, y = log10(RPKM))) +
  geom_boxplot(color = "black",
               position = position_dodge(width = 0.8), size = 0.8) +
  geom_jitter(aes(color = Type),
            alpha = 0.9, size = 2,
            position = position_jitterdodge(
        jitter.width = 0.5,
        dodge.width = 0.8)) +
  scale_color_manual(values = jitter_colors) +
  #get rid of box fill
  scale_fill_manual(values = c("transparent", "transparent", "transparent", "transparent", "transparent"), guide = "none") +
  guides(fill = "none") +
  facet_grid(subject ~ single_closest_ref, 
             scale = "free_y",
             labeller = labeller(single_closest_ref = species_labels)) +
  labs(
    title = "RPKM for unique dsrAB genes by sample location across subjects and species",
    x = "Sample Location",
    y = "log10(RPKM)",
    color = "Sample Location"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 10),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 60, hjust = 1),
    plot.title = element_text(size = 20, hjust = 0.5),
    plot.subtitle = element_text(size = 16, hjust = 0.5)
  )
ggsave("dsrAB_RPKM.png",
       width = 16, height = 14, units = "in",   # increase for larger output
       dpi = 600)
```


##fix below 
# individual plots
## subject 1
```{r}
#facet by species
samples_unique_genes_with_names %>%
  filter(subject == 1) %>%
  mutate(Type = factor(Type, levels = all_types)) %>%
  ggplot(aes(x = Type, y = log10(RPKM))) +
  geom_boxplot(color = "black",
               position = position_dodge(width = 0.8), size = 0.8) +
  geom_jitter(aes(color = Type),
            alpha = 0.9, size = 2,
            position = position_jitterdodge(
        jitter.width = 0.5,
        dodge.width = 0.8)) +
  scale_color_manual(values = jitter_colors, drop = FALSE) +
  scale_x_discrete(drop = FALSE) +
  #get rid of box fill
  scale_fill_manual(values = c("transparent", "transparent", "transparent", "transparent", "transparent"), guide = "none") +
  guides(fill = "none") +
  facet_wrap(~ single_closest_ref) +
  labs(
    title = "RPKM across capsule and stool samples by species",
    subtitle = "Subject 1",
    x = "Sample Type",
    y = "log10(RPKM)",
    color = "Type"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 14),
    legend.position = "none",
    axis.text.x = element_text(size = 12, angle = 60, hjust = 1),
    plot.title = element_text(size = 20, hjust = 0.5),
    plot.subtitle = element_text(size = 16, hjust = 0.5)
  )
#facet by type
samples_unique_genes_with_names %>%
  filter(subject == 1) %>%
  mutate(Type = factor(Type, levels = all_types),
    single_closest_ref = factor(single_closest_ref, levels = all_species)) %>%
  ggplot(aes(x = single_closest_ref, y = log10(RPKM))) +
  geom_boxplot(color = "black",
               position = position_dodge(width = 0.8), size = 0.8) +
  geom_jitter(aes(color = Type),
            alpha = 0.9, size = 2,
            position = position_jitterdodge(
        jitter.width = 0.5,
        dodge.width = 0.8)) +
  scale_color_manual(values = jitter_colors, drop = FALSE) +
  #scale_x_discrete(drop = FALSE) +
  #get rid of box fill
  scale_fill_manual(values = c("transparent", "transparent", "transparent", "transparent", "transparent"), guide = "none") +
  guides(fill = "none") +
  facet_wrap(~ Type, drop = FALSE) +
  labs(
    title = "RPKM across species by capsule and stool samples",
    subtitle = "Subject 1",
    x = "Closest Reference Species",
    y = "log10(RPKM)",
    color = "Type"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 14),
    legend.position = "none",
    axis.text.x = element_text(size = 12, angle = 60, hjust = 1),
    plot.title = element_text(size = 20, hjust = 0.5),
    plot.subtitle = element_text(size = 16, hjust = 0.5)
  )
```

