---
title: "clustering"
output: html_document
Author: "Elizabeth Daigle"
date: "16 March 2026"
---
# This rmd makes species-level clusters with ANI of >=90%, creates a plot showing cooccurrence of clusters in subjects, and creates a plot showing the percentage of SSRM-positive subjects and samples in which each cluster is found

# Read in packages and functions
```{r}
source(paste0(here::here(), "/1-Config/config.R"))
```

# Read in OTU table and taxa table
```{r}
taxa_table <- read_csv(paste0(clusterDir, "/taxa_table_0420.csv"))

otu_subject_matrix <- readRDS(paste0(dataDir, "/subject_OTU_matrix_0504.rds")) %>% 
  column_to_rownames(., var = "host subject id") %>%
  t() %>%
  as.matrix()
```
# Read in tree and root
```{r}
library(phyloseq)
ref_tree <- read.tree(paste0(raxmlDir, "/RAxML_originalLabelledTree_noBootstrap_no_2018_assemblies.newick"))
rooted_ref_tree <- root(ref_tree, c("Moorella_thermoacetica_Y72_copy_2__MooTher5"), r = TRUE)
TREE <- phy_tree(rooted_ref_tree)
```

# prep reference phyloseq object
```{r}
OTU = otu_table(otu_subject_matrix, taxa_are_rows = TRUE)

# format taxa_table
taxa_table <- taxa_table %>%
  column_to_rownames(var = "OTU") %>%
  select(-c("environmental_category","dsrab_type"))

taxa_table <- as.matrix(taxa_table)
# missing two taxa in the clusters_df (see code at bottom) so I am renaming them in taxa_table to match their names in the other phyloseq components. DslDes16 and DsvAlas8 have the same taxonomical classifications
rownames(taxa_table) <- dplyr::recode(rownames(taxa_table),
  "Candidatus_Desulfovibrio_trichonymphae__AB444072" = "Candidatus_Desulfovibrio_trichonymphae",
  "Desulfovibrio_desulfuricans_G20__DslDes16" = "Desulfovibrio_desulfuricans__DslDes16_DsvAlas8"
)
TAX = tax_table(taxa_table)

# make phyloseq object
phyloseq <- phyloseq(TREE, TAX, OTU)
```


# tip_glom at 0.15 and get list of OTUs that are in each cluster
```{r}
ps_tipglom <- tip_glom(phyloseq, h = 0.15)
# find which tips were retained after tip_glom
cluster_reps <- taxa_names(ps_tipglom)

# find the distances between tips on the tree so that each OTU can be assigned to its closest relative retained after tip_glom
tree_dist <- cophenetic(phy_tree(phyloseq))
```

# make df that gives the cluster representative for each of the 1256 taxa in the non tip-glommed ps object
```{r}
clusters_df <- data.frame(
  OTU = taxa_names(phyloseq),
  cluster = ifelse(taxa_names(phyloseq) %in% cluster_reps,
                   taxa_names(phyloseq),
                   sapply(taxa_names(phyloseq), function(tip) {
                     cluster_reps[which.min(tree_dist[tip, cluster_reps])]
                   }))
)

# read in final hits df 
hits_final <- readRDS(paste0(dataDir, "/hits_final_sample_prev.rds"))

# want to include only the OTUs that have hits 
names <- hits_final$single_closest_ref
hits_in_cluster_df <- clusters_df %>% filter(OTU %in% names) 

# testing in console
for (cluster in unique(hits_in_cluster_df$cluster)) {
  otus_in_cluster <- hits_in_cluster_df$OTU[hits_in_cluster_df$cluster == cluster]
  print(paste("Cluster:", cluster))
  print(otus_in_cluster)
  print(paste("Length:", length(otus_in_cluster)))
}

# # write one file per cluster
for (cluster in hits_in_cluster_df$cluster) {
  otus_in_cluster <- hits_in_cluster_df$OTU[hits_in_cluster_df$cluster == cluster]
  if (length(otus_in_cluster) > 1) {
    cluster_clean <- gsub("ö", "o", cluster)
    cluster_clean <- gsub("sp-", "sp.", cluster_clean)
    cluster_clean <- gsub("sk-", "sk.", cluster_clean)
    write_lines(otus_in_cluster,
                file = paste0(clusterDir, "/", cluster_clean, "_otus_in_cluster_015.txt"))
  }
}
```

```{r}
# ## missing two taxon at the end
# ## there are some OTUs from the reference tree fasta that are named differently than the reference taxonomy fasta (presumably they are two species/taxa that are very close)
# ps_object <- unique(hits_in_cluster_df$OTU)
# hits <- unique(hits_final$single_closest_ref)
# hits_not_ps_object <- setdiff(hits, ps_object)
# #moorella was removed and the NAs have "FALSE" for srm_present (these are not in the matrix because they were removed before making it, see dm_NCBI.rmd)
# # need to make sure DslDes16, DsvAlas8, and Candidatus Desulfovibrio triconymphae are kept 
# 
# # figure out where they are missing and change the names before creating ps object
# taxon <- "Candidatus_Desulfovibrio_trichonymphae"
# 
# # Check presence in each component
# taxon %in% phy_tree(TREE)$tip.label        # in tree?
# taxon %in% rownames(taxa_table)            # in TAX?
# taxon %in% rownames(otu_subject_matrix)    # in OTU?
# 
# taxon2 <- "Desulfovibrio_desulfuricans__DslDes16_DsvAlas8"
# 
# # Check presence in each component
# taxon2 %in% phy_tree(TREE)$tip.label        # in tree?
# taxon2 %in% rownames(taxa_table)            # in TAX?
# taxon2 %in% rownames(otu_subject_matrix)    # in OTU?
```




# rename the OTUs in the matrix to species-level based on the clustering 
```{r}
otu_subject_matrix_subset <- otu_subject_matrix %>% .[rowSums(.) > 0, ] 


before = rownames(otu_subject_matrix)
after = rownames(otu_subject_matrix_subset)
setdiff(before, after)
# lost these three OTUs (Polluted_aquifer_clone__UncS1623, Polluted_aquifer_clone_LGWK16__UncS1633, Desulfosporosinus_youngiae__DfpYoun5) because the only samples they were present in were removed from the subject df in dm_NCBI.Rmd

species_subject_matrix <- otu_subject_matrix[rownames(otu_subject_matrix) 
                                                 != "Moorella_thermoacetica_Y72_copy_2__MooTher5", ]

rownames(species_subject_matrix) <- hits_in_cluster_df$cluster[match(rownames(species_subject_matrix), 
                                                                     hits_in_cluster_df$OTU)]
row_names <- rownames(otu_subject_matrix_subset)
# seeing if the OTUs from the matrix match the OTUs from the clustering
setdiff(row_names, hits_in_cluster_df$OTU)
# only moorella thermoacetica is missing which is what I expect since that is used for rooting the tree
setdiff(hits_in_cluster_df$OTU, row_names)
# again, the two OTUs lost are only found in samples not in the subject matrix; clustering is based off the taxa table, not the otu matrix

# double check the row sums are the same before and after renaming -> yes!
# Bilophila cluster -> 825 subjects with a hit
rowSums(otu_subject_matrix_subset[c("Bilophila_wadsworthia__BilWads6", "Bilophila_sp._4_1_30__BilSpec3"), ])
rowSums(species_subject_matrix[grep("Bilophila_wadsworthia", rownames(species_subject_matrix)), ])
# D. piger -> 333 subjects with a hit
rowSums(otu_subject_matrix_subset[c("Desulfovibrio_piger__DslPige2","Desulfovibrio_piger__DslPige3", "Desulfovibrio_piger__DslPige4"), ])
rowSums(species_subject_matrix[grep("Desulfovibrio_piger", rownames(species_subject_matrix)), ])
# desulfovibrio cluster -> 165 subjects with a hit
rowSums(otu_subject_matrix_subset[c("uncultured_bacterium__EF645664","uncultured_bacterium__EF645667","uncultured_bacterium__EF645665", "Desulfovibrio_africanus__DslAfri3", "Desulfovibrio_simplex__DslSimpl", "Desulfovibrio_desulfuricans_subsp._desulfuricans_str._ATCC_27774__CP001358", "Desulfovibrio_desulfuricans__DslDes17"), ])
rowSums(species_subject_matrix[grep("uncultured_bacterium__EF645667", rownames(species_subject_matrix)), ])
```
## collapse the rows by cluster and make the cooccurrence network
```{r}
groups <- gsub("\\.[1-6]$", "", rownames(species_subject_matrix))
species_subject_matrix_by_cluster <- apply(species_subject_matrix, 2, function(col) {
  tapply(col, groups, max)
})

rowSums(species_subject_matrix_by_cluster[1:23, ])
# sums are different now since each cluster has a value of 1 (rather than a value of 1 for each species in the cluster)

# 610 subjects with either bilophila species, as expected
bil_cluster_check <- otu_subject_matrix[c("Bilophila_wadsworthia__BilWads6", "Bilophila_sp._4_1_30__BilSpec3"), colSums(otu_subject_matrix[c("Bilophila_wadsworthia__BilWads6", "Bilophila_sp._4_1_30__BilSpec3"), ]) >= 1]
```
# cooccur with cutoff of 20 subjects
```{r}
cooccur.subjectCluster <- cooccur(mat=species_subject_matrix_by_cluster, spp_names = T)
edgeList <- cooccur.subjectCluster$results[, c("sp1_name", "sp2_name", "obs_cooccur")]
edgeList <- subset(edgeList, obs_cooccur > 20)

igraph_net <- graph_from_data_frame(d = edgeList, directed = F)
E(igraph_net)$width <- E(igraph_net)$obs_cooccur/8
V(igraph_net)$label.size <- 0
deg <- igraph::degree(igraph_net, mode="all")
V(igraph_net)$size <- 1
species_labels <- c(
  "Bilophila_wadsworthia__BilWads6" = expression(italic("Bilophila wadsworthia")),
  "Desulfovibrio_piger__DslPige3" = expression(italic("Desulfovibrio piger")),
  "uncultured_bacterium__EF645667" = expression(paste("uncultured ", italic("Desulfovibrio"), " species")),
  "Desulfitibacter_alkalitolerans__DfiAlka2" = expression(italic("Desulfitibacter alkalitolerans")),
  "Anaerobic_bacterium_sk.prop8__Bv2Prop2" = "uncultured Firmicute from industrial environment AY751286"
)
V(igraph_net)$label <- species_labels[V(igraph_net)$name]

# add colors by supercluster
taxa_table_df <- read_csv(paste0(clusterDir, "/taxa_table_0420.csv")) 
# supercluster identity
desulfo_family <- taxa_table_df %>% filter(supercluster == "Deltaproteobacteria supercluster")
firmicutes <- taxa_table_df %>% filter(supercluster == "FACA group sensu lato")
environ <- taxa_table_df %>% filter(supercluster == "Environmental supercluster 1")
archaeoglobus <- taxa_table_df %>% filter(supercluster == "Archaeoglobus cluster")

#colors 
firmicute_orange <- brewer.pal(2, "Set2")[2]
desulfovibrionaceae_purple <- brewer.pal(10, "Set3")[10]
colors <- brewer.pal(name = "Set2", 8)

V(igraph_net)$label.color <- "black"  # default
V(igraph_net)[name %in% desulfo_family$OTU]$label.color <- desulfovibrionaceae_purple
V(igraph_net)[name %in% firmicutes$OTU]$label.color <- firmicute_orange
V(igraph_net)[name %in% environ$OTU]$label.color <- colors[1]
V(igraph_net)[name %in% archaeoglobus$OTU]$label.color <- colors[3]

# E(igraph_net)$color <- firmicute_orange  # default single line color
# E(igraph_net)[.inc(V(igraph_net)[name %in% desulfo])]$color <- desulfovibrionaceae_purple

radian.rescale <- function(x, start=0, direction=1) {
  c.rotate <- function(x) (x + start) %% (2 * pi) * direction
  c.rotate(scales::rescale(x, c(0, 2 * pi), range(x)))
}
lab.locs <- radian.rescale(x=1:vcount(igraph_net), direction=-1, start=0)
#setting distances from vertices for labels
lab.locs[V(igraph_net)$name == "Anaerobic_bacterium_sk.prop8__Bv2Prop2"] <- -pi/20
lab.locs[V(igraph_net)$name == "uncultured_bacterium__EF645667"] <- pi/3
V(igraph_net)["Anaerobic_bacterium_sk.prop8__Bv2Prop2"]$label.dist <- 10
V(igraph_net)["Bilophila_wadsworthia__BilWads6"]$label.dist <- 2
V(igraph_net)["Desulfovibrio_piger__DslPige3"]$label.dist <- 7
V(igraph_net)["uncultured_bacterium__EF645667"]$label.dist <- 2
V(igraph_net)["Desulfitibacter_alkalitolerans__DfiAlka2"]$label.dist <- 2

l <- layout_in_circle(igraph_net)

plot(igraph_net, layout = l, edge.curved=.3, vertex.label = NA, edge.alpha = 0.5, vertex.color = "black", vertex.frame.color = "black")

jpeg("cooccurrence_20subj_cutoff.jpeg", width = 9, height = 4.5, units = "in", res = 300)
plot(igraph_net, layout = l, edge.curved=.3, vertex.color = "black", vertex.frame.color = "black", vertex.label.cex = 0.7, vertex.label.color = V(igraph_net)$label.color, vertex.label.degree = lab.locs, edge.alpha = 0.5)
dev.off()
```

# some double checking
```{r}
# there are indeed 77 co-occurrences of bilophila and d. piger (matches edgelist) 
edgelist_check <- species_subject_matrix_by_cluster[c("Bilophila_wadsworthia__BilWads6", "Desulfovibrio_piger__DslPige3"), colSums(species_subject_matrix_by_cluster[c("Bilophila_wadsworthia__BilWads6", "Desulfovibrio_piger__DslPige3"), ]) == 2]
```

# cooccur with cutoff of 15 subjects
```{r}
cooccur.subjectCluster <- cooccur(mat=species_subject_matrix_by_cluster, spp_names = T)
edgeList <- cooccur.subjectCluster$results[, c("sp1_name", "sp2_name", "obs_cooccur")]
edgeList <- subset(edgeList, obs_cooccur > 15)

igraph_net <- graph_from_data_frame(d = edgeList, directed = F)
E(igraph_net)$width <- E(igraph_net)$obs_cooccur/8
V(igraph_net)$label.size <- 0
deg <- igraph::degree(igraph_net, mode="all")
V(igraph_net)$size <- 1
species_labels <- c(
  "Bilophila_wadsworthia__BilWads6" = expression(italic("Bilophila wadsworthia")),
  "Desulfovibrio_piger__DslPige3" = expression(italic("Desulfovibrio piger")),
  "uncultured_bacterium__EF645667" = expression(paste("uncultured ", italic("Desulfovibrio"), " species")),
  "Desulfitibacter_alkalitolerans__DfiAlka2" = expression(italic("Desulfitibacter alkalitolerans")),
  "Anaerobic_bacterium_sk.prop8__Bv2Prop2" = "uncultured Firmicute from industrial environment AY751286",
  "Gordonibacter_pamelaeae__GrdPame5" = expression(italic("Gordonibacter pamelaeae")))
V(igraph_net)$label <- species_labels[V(igraph_net)$name]

# add colors by supercluster
taxa_table_df <- read_csv(paste0(clusterDir, "/taxa_table_0420.csv")) 
# supercluster identity
desulfo_family <- taxa_table_df %>% filter(supercluster == "Deltaproteobacteria supercluster")
firmicutes <- taxa_table_df %>% filter(supercluster == "FACA group sensu lato")
environ <- taxa_table_df %>% filter(supercluster == "Environmental supercluster 1")
archaeoglobus <- taxa_table_df %>% filter(supercluster == "Archaeoglobus cluster")

#colors 
firmicute_orange <- brewer.pal(2, "Set2")[2]
desulfovibrionaceae_purple <- brewer.pal(10, "Set3")[10]
colors <- brewer.pal(name = "Set2", 8)

V(igraph_net)$label.color <- "black"  # default
V(igraph_net)[name %in% desulfo_family$OTU]$label.color <- desulfovibrionaceae_purple
V(igraph_net)[name %in% firmicutes$OTU]$label.color <- firmicute_orange
V(igraph_net)[name %in% environ$OTU]$label.color <- colors[1]
V(igraph_net)[name %in% archaeoglobus$OTU]$label.color <- colors[3]

# E(igraph_net)$color <- firmicute_orange  # default single line color
# E(igraph_net)[.inc(V(igraph_net)[name %in% desulfo])]$color <- desulfovibrionaceae_purple

radian.rescale <- function(x, start=0, direction=1) {
  c.rotate <- function(x) (x + start) %% (2 * pi) * direction
  c.rotate(scales::rescale(x, c(0, 2 * pi), range(x)))
}
lab.locs <- radian.rescale(x=1:vcount(igraph_net), direction=-1, start=0)
#setting distances from vertices for labels
lab.locs[V(igraph_net)$name == "Anaerobic_bacterium_sk.prop8__Bv2Prop2"] <- -pi/20
lab.locs[V(igraph_net)$name == "uncultured_bacterium__EF645667"] <- pi/2.5
lab.locs[V(igraph_net)$name == "Desulfovibrio_piger__DslPige3"] <- -pi/2
lab.locs[V(igraph_net)$name == "Gordonibacter_pamelaeae__GrdPame5"] <- pi/2
# 
V(igraph_net)["Anaerobic_bacterium_sk.prop8__Bv2Prop2"]$label.dist <- 17
V(igraph_net)["Bilophila_wadsworthia__BilWads6"]$label.dist <- 2
V(igraph_net)["Desulfovibrio_piger__DslPige3"]$label.dist <- 2
V(igraph_net)["uncultured_bacterium__EF645667"]$label.dist <- 2
V(igraph_net)["Desulfitibacter_alkalitolerans__DfiAlka2"]$label.dist <- -3
V(igraph_net)["Gordonibacter_pamelaeae__GrdPame5"]$label.dist <- 1


l <- layout_in_circle(igraph_net)

plot(igraph_net, layout = l, edge.curved=.3, vertex.label = NA, edge.alpha = 0.5, vertex.color = "black", vertex.frame.color = "black")

jpeg("cooccurrence_15subj_cutoff.jpeg", width = 9, height = 4.5, units = "in", res = 300)
plot(igraph_net, layout = l, edge.curved=.3, vertex.color = "black", vertex.frame.color = "black", vertex.label.cex = 0.7, vertex.label.color = V(igraph_net)$label.color, vertex.label.degree = lab.locs, edge.alpha = 0.5)
dev.off()
```

# cooccur with cutoff of 10 subjects
```{r}
cooccur.subjectCluster <- cooccur(mat=species_subject_matrix_by_cluster, spp_names = T)
edgeList <- cooccur.subjectCluster$results[, c("sp1_name", "sp2_name", "obs_cooccur")]
edgeList <- subset(edgeList, obs_cooccur > 10)

igraph_net <- graph_from_data_frame(d = edgeList, directed = F)
E(igraph_net)$width <- E(igraph_net)$obs_cooccur/8
V(igraph_net)$label.size <- 0
deg <- igraph::degree(igraph_net, mode="all")
V(igraph_net)$size <- 1
species_labels <- c(
  "Bilophila_wadsworthia__BilWads6" = expression(italic("Bilophila wadsworthia")),
  "Desulfovibrio_piger__DslPige3" = expression(italic("Desulfovibrio piger")),
  "uncultured_bacterium__EF645667" = expression(paste("uncultured ", italic("Desulfovibrio"), " species")),
  "Desulfitibacter_alkalitolerans__DfiAlka2" = expression(italic("Desulfitibacter alkalitolerans")),
  "Anaerobic_bacterium_sk.prop8__Bv2Prop2" = "uncultured Firmicute from industrial environment AY751286",
  "Desulfovibrio_sp._6_1_46AFAA__DsvSp231" = expression(paste(italic("Desulfovibrio"), " species")),
  "Desulfovibrio_vulgaris__DQ826729" = expression(italic("Desulfovibrio vulgaris")),
  "Gordonibacter_pamelaeae__GrdPame5" = expression(italic("Gordonibacter pamelaeae")))

V(igraph_net)$label <- species_labels[V(igraph_net)$name]

# add colors by supercluster
taxa_table_df <- read_csv(paste0(clusterDir, "/taxa_table_0420.csv")) 
# supercluster identity
desulfo_family <- taxa_table_df %>% filter(supercluster == "Deltaproteobacteria supercluster")
firmicutes <- taxa_table_df %>% filter(supercluster == "FACA group sensu lato")
environ <- taxa_table_df %>% filter(supercluster == "Environmental supercluster 1")
archaeoglobus <- taxa_table_df %>% filter(supercluster == "Archaeoglobus cluster")

#colors 
firmicute_orange <- brewer.pal(2, "Set2")[2]
desulfovibrionaceae_purple <- brewer.pal(10, "Set3")[10]
colors <- brewer.pal(name = "Set2", 8)

V(igraph_net)$label.color <- "black"  # default
V(igraph_net)[name %in% desulfo_family$OTU]$label.color <- desulfovibrionaceae_purple
V(igraph_net)[name %in% firmicutes$OTU]$label.color <- firmicute_orange
V(igraph_net)[name %in% environ$OTU]$label.color <- colors[1]
V(igraph_net)[name %in% archaeoglobus$OTU]$label.color <- colors[3]

# E(igraph_net)$color <- firmicute_orange  # default single line color
# E(igraph_net)[.inc(V(igraph_net)[name %in% desulfo])]$color <- desulfovibrionaceae_purple

radian.rescale <- function(x, start=0, direction=1) {
  c.rotate <- function(x) (x + start) %% (2 * pi) * direction
  c.rotate(scales::rescale(x, c(0, 2 * pi), range(x)))
}
lab.locs <- radian.rescale(x=1:vcount(igraph_net), direction=-1, start=0)
#setting distances from vertices for labels
lab.locs[V(igraph_net)$name == "Anaerobic_bacterium_sk.prop8__Bv2Prop2"] <- -pi/25
lab.locs[V(igraph_net)$name == "uncultured_bacterium__EF645667"] <- pi/3
lab.locs[V(igraph_net)$name == "Desulfovibrio_piger__DslPige3"] <- -pi/2
lab.locs[V(igraph_net)$name == "Gordonibacter_pamelaeae__GrdPame5"] <- pi/2
lab.locs[V(igraph_net)$name == "Desulfitibacter_alkalitolerans__DfiAlka2"] <- -pi/2
# 
V(igraph_net)["Anaerobic_bacterium_sk.prop8__Bv2Prop2"]$label.dist <- 17
V(igraph_net)["Bilophila_wadsworthia__BilWads6"]$label.dist <- 2
V(igraph_net)["Desulfovibrio_piger__DslPige3"]$label.dist <- 2
V(igraph_net)["uncultured_bacterium__EF645667"]$label.dist <- 2
V(igraph_net)["Desulfitibacter_alkalitolerans__DfiAlka2"]$label.dist <- 2
V(igraph_net)["Gordonibacter_pamelaeae__GrdPame5"]$label.dist <- 2
V(igraph_net)["Desulfovibrio_vulgaris__DQ826729"]$label.dist <- 2
V(igraph_net)["Desulfovibrio_sp._6_1_46AFAA__DsvSp231"]$label.dist <- 2

l <- layout_in_circle(igraph_net)

plot(igraph_net, layout = l, edge.curved=.3, vertex.label = NA, edge.alpha = 0.5, vertex.color = "black", vertex.frame.color = "black")

jpeg("cooccurrence_10subj_cutoff.jpeg", width = 9, height = 4.5, units = "in", res = 300)
plot(igraph_net, layout = l, edge.curved=.3, vertex.color = "black", vertex.frame.color = "black", vertex.label.cex = 0.7, vertex.label.color = V(igraph_net)$label.color, vertex.label.degree = lab.locs, edge.alpha = 0.5)
dev.off()
```










## FOR SAMPLE LEVEL PREVALENCE CALCULATIONS
# read in the sample_OTU_matrix 
# rename the OTUs in the matrix to species-level based on the clustering 
```{r}
otu_sample_matrix <- readRDS(file = paste0(dataDir, "/sample_OTU_matrix_0506.rds")) %>%
  column_to_rownames(., var = "sample_id") %>%
  t() %>%
  as.matrix()

species_sample_matrix <- otu_sample_matrix[rownames(otu_sample_matrix) 
                                                 != "Moorella_thermoacetica_Y72_copy_2__MooTher5", ]

rownames(species_sample_matrix) <- hits_in_cluster_df$cluster[match(rownames(species_sample_matrix), 
                                                                     hits_in_cluster_df$OTU)]
otu_sample_matrix_subset <- otu_sample_matrix %>% .[rowSums(.) > 0, ]
row_names <- rownames(otu_sample_matrix_subset)
# seeing if the OTUs from the matrix match the OTUs from the clustering
setdiff(row_names, hits_in_cluster_df$OTU)
# only moorella thermoacetica is missing which is what I expect since that is used for rooting the tree
setdiff(hits_in_cluster_df$OTU, row_names)
# nothing missing now since all of the samples are included in this matrix

# double check the row sums are the same before and after renaming -> yes!
# Bilophila cluster -> 1014 samples with a hit
rowSums(otu_sample_matrix_subset[c("Bilophila_wadsworthia__BilWads6", "Bilophila_sp._4_1_30__BilSpec3"), ])
rowSums(species_sample_matrix[grep("Bilophila_wadsworthia", rownames(species_sample_matrix)), ])
# D. piger -> 404 samples with a hit
rowSums(otu_sample_matrix_subset[c("Desulfovibrio_piger__DslPige2","Desulfovibrio_piger__DslPige3", "Desulfovibrio_piger__DslPige4"), ])
rowSums(species_sample_matrix[grep("Desulfovibrio_piger", rownames(species_sample_matrix)), ])
# desulfovibrio cluster -> 207 samples with a hit
rowSums(otu_sample_matrix_subset[c("uncultured_bacterium__EF645664","uncultured_bacterium__EF645667","uncultured_bacterium__EF645665", "Desulfovibrio_africanus__DslAfri3", "Desulfovibrio_simplex__DslSimpl", "Desulfovibrio_desulfuricans_subsp._desulfuricans_str._ATCC_27774__CP001358", "Desulfovibrio_desulfuricans__DslDes17"), ])
rowSums(species_sample_matrix[grep("uncultured_bacterium__EF645667", rownames(species_sample_matrix)), ])
```
## collapse the rows by cluster
```{r}
groups_sample <- gsub("\\.[1-6]$", "", rownames(species_sample_matrix))
species_sample_matrix_by_cluster <- apply(species_sample_matrix, 2, function(col) {
  tapply(col, groups_sample, max)
})

rowSums(species_sample_matrix_by_cluster[1:23, ])
# sums are different now since each cluster has a value of 1 (rather than a value of 1 for each species in the cluster)

#this looks right! 757 samples where either of the bilophila are present
bil_cluster_check <- otu_sample_matrix[c("Bilophila_wadsworthia__BilWads6", "Bilophila_sp._4_1_30__BilSpec3"), colSums(otu_sample_matrix[c("Bilophila_wadsworthia__BilWads6", "Bilophila_sp._4_1_30__BilSpec3"), ]) >= 1]

```

# prevalence of clusters across SSRM-positive subjects and samples
```{r}
total_samples = ncol(species_sample_matrix_by_cluster)
total_subjects = ncol(species_subject_matrix_by_cluster)

n_subj = rowSums(species_subject_matrix_by_cluster[1:23, ])
n_samples = rowSums(species_sample_matrix_by_cluster[1:23, ])
cluster_prevalence <- data.frame(
  n_subj = n_subj,
  # 960 is the number of subjects in the matrix
  subj_prev = (n_subj/total_subjects),
  n_samples = n_samples,
  sample_prev = (n_samples/total_samples)) %>%
  rownames_to_column(var = "cluster")

taxa_table_df <- read_csv(paste0(clusterDir, "/taxa_table_0420.csv")) 


#make larger groups for each cluster
cluster_groups_fig <- taxa_table_df %>% filter(OTU %in% cluster_prevalence$cluster | str_detect(OTU, "trich|DslDes16"))

desulfo_family <- cluster_groups_fig %>% filter(supercluster == "Deltaproteobacteria supercluster")
firmicutes <- cluster_groups_fig %>% filter(supercluster == "FACA group sensu lato")
environ <- cluster_groups_fig %>% filter(supercluster == "Environmental supercluster 1")
archaeoglobus <- cluster_groups_fig %>% filter(supercluster == "Archaeoglobus cluster")

cluster_prevalence_long <- cluster_prevalence %>%
  select(-c("n_subj", "n_samples")) %>%
  pivot_longer(c("subj_prev", "sample_prev")) %>%
  mutate(group = case_when(
    cluster %in% desulfo_family$OTU | str_detect(cluster, "Candidatus_Desulfovibrio_trichonymphae|DslDes16") ~ "Desulfovibrionaceae family",
    cluster %in% firmicutes$OTU ~ "Firmicutes group",
    cluster %in% environ$OTU ~ "Environmental supercluster 1",
    cluster %in% archaeoglobus$OTU ~ "Archaeoglobus supercluster")) %>%
  mutate(cluster_label = case_when(
    cluster == "Anaerobic_bacterium_EtOH8__Bv2EtOH2" ~ "uncultured Firmicute from industrial environment AY751284",
    cluster == "Anaerobic_bacterium_sk.prop8__Bv2Prop2" ~ "uncultured Firmicute from industrial environment AY751286",
    cluster == "Archaeoglobus_sulfaticallidus__ArgSulfa" ~ "*Archaeoglobus sulfaticallidus*",
    cluster == "Arctic_sediment_clone_AH5_5__AtcSed67" ~ "uncultured Firmicute from arctic sediment",
    cluster == "Bilophila_wadsworthia__BilWads6" ~ "*Bilophila wadsworthia*",
    cluster == "Bioreactor_clone__UncS2029" ~ "uncultured Firmicute from bioreactor",
    cluster == "Candidatus_Desulfovibrio_trichonymphae" ~ "*Candidatus Desulfovibrio trichonymphae*",
    cluster == "Desulfitibacter_alkalitolerans__DfiAlka2" ~ "*Desulfitibacter alkalitolerans*",
    cluster == "Desulfosporosinus_youngiae__DfpYoun5" ~ "*Desulfosporosinus youngiae*",
    cluster == "Desulfotomaculum_hydrothermale__DscHydr2" ~ "*Desulfotomaculum hydrothermale*",
    cluster == "Desulfovibrio_carbinoliphilus__JQ031012" ~ "*Desulfovibrio carbinoliphilus*",
    cluster == "Desulfovibrio_desulfuricans__DQ092635" ~ "*Desulfovibrio desulfuricans* DQ092635",
    cluster == "Desulfovibrio_desulfuricans__DslDes16_DsvAlas8" ~ "*Desulfovibrio desulfuricans* NC_007519",
    cluster == "Desulfovibrio_piger__DslPige3" ~ "*Desulfovibrio piger*",
    cluster == "Desulfovibrio_sp._6_1_46AFAA__DsvSp231" ~ "*Desulfovibrio* species ACWM01000066",
    cluster == "Desulfovibrio_vulgaris__DQ826729" ~ "*Desulfovibrio vulgaris*",
    cluster == "Desulfurispora_thermophila__DprTher2" ~ "*Desulfurispora thermophila*",
    cluster == "Gordonibacter_pamelaeae__GrdPame5" ~ "*Gordonibacter pamelaeae*",
    cluster == "MSW_digester_clone__Unc01cxc" ~ "uncultured Firmicute from digester",
    cluster == "Polluted_aquifer_clone_LGWK16__UncS1633" ~ "uncultured Firmicute from aquifer",
    cluster == "Rice_field_soil_clone_Bu-23__JN615148" ~ "uncultured bacterium from rice field",
    cluster == "Thermodesulfobium_narugense__CP002690" ~ "*Thermodesulfobium narugense*",
    cluster == "uncultured_bacterium__EF645667" ~ "uncultured *Desulfovibrio* species",
    TRUE ~ cluster
  ))

cluster_prevalence_long$group <- factor(cluster_prevalence_long$group, 
                                         levels = c("Desulfovibrionaceae family",
                                                    "Firmicutes group",
                                                    "Environmental supercluster 1",
                                                    "Archaeoglobus supercluster"))
cluster_prevalence_long$name <- factor(cluster_prevalence_long$name, 
                                         levels = c("subj_prev", "sample_prev"))

colors <- brewer.pal(name = "Set2", 8)
desulfovibrionaceae_purple <- brewer.pal(10, "Set3")[10]
breaks = c(0, 0.1, 1, 10, 100)
#minor_breaks = c()
(plot <- ggplot(cluster_prevalence_long, aes(y = reorder(cluster_label, value), x = value*100, color = group)) +
  geom_point(size = 2) +
  scale_color_manual(values = c(desulfovibrionaceae_purple, colors[2], colors[1], colors[3])) +
  facet_grid(group~name, scales = "free_y", space = "free") +
  theme_bw() +
  #scale_fill_manual(values = c(colors[3], colors[4])) +
  guides(fill = "none", color = "none") +
  theme(axis.title = element_blank(), strip.background = element_blank(), strip.text.x = element_blank(), strip.text.y.right = element_text(angle = 0, size = 8), axis.text.y= element_markdown(size = 8), axis.text.x= element_text(size = 8)) +
    scale_x_log10(breaks = breaks, limits = c(NA,100)))

ggsave(plot = plot, width = 8, height = 5, file = "percentages_barchart_0506.jpeg")


```

# for results, percent of subjects with desulfo and firmicutes
```{r}
# find the number of columns (subjects) in the matrix where there is at least one hit to a firmicutes cluster
subj_num_firm <- sum(colSums(species_subject_matrix_by_cluster[rownames(species_subject_matrix_by_cluster) %in% firmicutes$OTU, ]) >= 1)
subj_num_firm/960

# find the number of columns (subjects) in the matrix where there is at least one hit to a desulfovibrionaceae cluster
subj_num_des <- sum(colSums(species_subject_matrix_by_cluster[rownames(species_subject_matrix_by_cluster) %in% desulfo_family$OTU, ]) >= 1)
subj_num_des/960


desulfo_family_no_bil <- cluster_groups_fig %>% filter(supercluster == "Deltaproteobacteria supercluster") %>%
  filter(!OTU == "Bilophila_wadsworthia__BilWads6")
# excluding bilophila prevalence
subj_num_des_no_bil <- sum(colSums(species_subject_matrix_by_cluster[rownames(species_subject_matrix_by_cluster) %in% desulfo_family_no_bil$OTU, ]) >= 1)
subj_num_des_no_bil/960

```

